---
title: "Machine Learning Lab"
author: "Yukun Jiao"
date: "2025-09-24"
toc: false
format: pdf
knitr: 
  opts_chunk:
    fig.align: center
editor: 
  markdown: 
    wrap: 72
---

# Lab 5

## Part 1:

### 1.

```{r}
#| warning: false
#| message: false

library(data.table)
library(dplyr)
library(ggplot2)
library(slam)         # New: needed for perplexity calc
library(quanteda)     # New: to process text
library(topicmodels)  # New: to fit topic models
library(word2vec)     # New: to fit word embeddings
library(tidyr)
library(purrr)
library(knitr)
library(kableExtra)
```


```{r}
fb_congress_data <- fread("~/Documents/ML/Labs/Lab5/fb-congress-data3.csv")
dim(fb_congress_data)
colnames(fb_congress_data)
```

The dataset contains 6752 rows and 4 columns.
The variable names are: `doc_id`, `screen_name`, `party`, and `message`.

### 2
```{r}
posts_corpus <- corpus(x = fb_congress_data,
                       text_field = "message",
                       meta = list("screen_name", "party"),
                       docid_field = 'doc_id')

# print(posts_corpus)

# Tokenize & clean from particular types of words
mytokens <- tokens(x = posts_corpus, 
                   remove_punct = TRUE, 
                   remove_numbers = TRUE, 
                   remove_symbols = TRUE,
                   remove_url = TRUE,
                   padding = FALSE)
# print(mytokens)

# Remove English stop words
mytokens <-   tokens_remove(x = mytokens,
                            stopwords("en"), 
                            padding = FALSE)
# Remove @'s
mytokens <-   tokens_select(x = mytokens,
                            selection = 'remove',
                            valuetype = 'glob', 
                            pattern = '@', 
                            padding = FALSE)

# Make tokens lowercase
mytokens <- tokens_tolower(x = mytokens)

# How does it look now?
mytokens[1:3]
```
I printed the first three texts for inspection, and I can confirm that the data has been successfully processed into tokens.

```{r}
# Create document term matrix
dtm <- dfm(x = mytokens)

# Exclude words with too low frequency
dtm <- dfm_trim(dtm, min_termfreq = 5)

# Exclude documents with too low frequency
rowsums <- rowSums(dtm)
keep_ids <- which(rowsums>=10)
dtm <- dtm[keep_ids,]

# Report final dimensionality of data set
dim(dtm)
```
The final document-term matrix has dimensions of 5,748 rows (documents) and 5,484 columns (terms).

### 3. LDA

```{r}
# Fit LDA (package: topicmodels)
# - 30 topics
# - Estimation algorithm: Gibbs
# - 1000 iterations
# - Set seed for replicability
# - Print iter every 1000th (I need the "quiet" model)
set.seed(1)
K <- 50
mylda <- LDA(x = dtm, 
             k = K, 
             method="Gibbs",
             control=list(iter = 1000, 
                          seed = 1, 
                          verbose = 0))
```

### 4.
```{r}

# Inspect top 15 from each topics
get_terms(mylda, 15)

# get_terms(mylda, 15)[, c(3, 4, 16, 30, 36)]

# To get a more granular view, extract the probabilities 
mylda_posterior <- topicmodels::posterior(object = mylda)
topic_distr_over_words <- mylda_posterior$terms
topic_distr_over_words_dt <- data.table(topic=1:K, 
                                        topic_distr_over_words)
topic_distr_over_words_dt <- melt.data.table(topic_distr_over_words_dt, # If you use base R, you can use reshape's "melt()" function.
                                             id.vars = 'topic')


# Tidyr/Dplyr way of extracting top 15 rows by group
top10per_topic <- topic_distr_over_words_dt %>% 
  group_by(topic) %>% 
  slice_max(order_by = value, n = 15)

# data.table way of extracting top 10 rows by group
topic_distr_over_words_dt <- topic_distr_over_words_dt[order(value,decreasing = T)]
top10per_topic <- topic_distr_over_words_dt[,head(.SD,10),by='topic']

topic_name <- c("national security", "drug", "education", "agriculture", "healthcare system")
# Plot probability over words for a few topics.
ggplot(top10per_topic[topic %in% c(3, 4, 16, 30, 36)],aes(y=factor(variable),x=value)) + 
  geom_bar(stat = 'identity', position = 'dodge') + 
  facet_wrap(~factor(topic,
                     levels = c(3, 4, 16, 30, 36),
                     labels = c("national security", "drug", "education", "agriculture", "healthcare system")
                     ),
             scales = 'free')
```

After examining the top 15 words for all topics, I selected topics 3, 4, 16, 30, and 36 and provided appropriate labels for each.

Overall, most topics are clear and meaningful, but there are some "junk" topics. For example, topic 50 do not convey any meaningful information and mainly consist of irrelevant words.

### 5
```{r}
# To validate labeling; identify documents with highest proportion on any given topic
doc_topic_proportions <- mylda_posterior$topics
# dim(doc_topic_proportions)
# head(doc_topic_proportions)
doc_topic_proportions_dt <- data.table(doc_id = mylda@documents,
                                       doc_topic_proportions)
# here I choose topic 3 and topic 36

colnames(doc_topic_proportions_dt)[2:ncol(doc_topic_proportions_dt)] <- paste0('Topic',colnames(doc_topic_proportions_dt)[2:ncol(doc_topic_proportions_dt)]) # Assign topics as column names

topic3doc <- doc_topic_proportions_dt[order(Topic3,decreasing = T)][1:3,]
topic36doc <- doc_topic_proportions_dt[order(Topic36,decreasing = T)][1:3,]


# Select rows from original data...
# national security
cat("Top 3 documents for topic 3 are:", topic3doc$doc_id)

# fb_congress_data[doc_id %in% topic3doc$doc_id,]$message

# obamacare
cat("Top 3 documents for topic 36 are:", topic36doc$doc_id)

# fb_congress_data[doc_id %in% topic36doc$doc_id,]$message
```
From the topics I labeled, I chose topic 3 and topic 36, which correspond to “national security” and “healthcare system”, respectively.

For each of these two topics, I reported the three documents with the highest topic proportions. After reading them qualitatively, I found that they indeed support the themes I assigned.

### 6.
```{r}
set.seed(1)
K <- 3
mylda_3 <- LDA(x = dtm, 
             k = K, 
             method="Gibbs",
             control=list(iter = 1000, 
                          seed = 1, 
                          verbose = 0))

# Inspect top 10 from each topics
get_terms(mylda_3, 10)

# To get a more granular view, extract the probabilities 
mylda_3_posterior <- topicmodels::posterior(object = mylda_3)
topic_distr_over_words <- mylda_3_posterior$terms
topic_distr_over_words_dt <- data.table(topic=1:K, 
                                        topic_distr_over_words)
topic_distr_over_words_dt <- melt.data.table(topic_distr_over_words_dt, # If you use base R, you can use reshape's "melt()" function.
                                             id.vars = 'topic')


# Tidyr/Dplyr way of extracting top 15 rows by group
top10per_topic <- topic_distr_over_words_dt %>% 
  group_by(topic) %>% 
  slice_max(order_by = value, n = 15)

# data.table way of extracting top 10 rows by group
topic_distr_over_words_dt <- topic_distr_over_words_dt[order(value,decreasing = T)]
top10per_topic <- topic_distr_over_words_dt[,head(.SD,10),by='topic']

# Plot probability over words for a few topics.
ggplot(top10per_topic[topic %in% 1:3],aes(y=factor(variable),x=value)) + 
  geom_bar(stat = 'identity', position = 'dodge') + 
  facet_wrap(~factor(topic,
                     labels = c("nation", "politics", "family")
                     ),
             scales = 'free')

# To validate labeling; identify documents with highest proportion on any given topic
doc_topic_proportions <- mylda_3_posterior$topics
dim(doc_topic_proportions)
head(doc_topic_proportions)
doc_topic_proportions_dt <- data.table(doc_id = mylda_3@documents,
                                       doc_topic_proportions)
colnames(doc_topic_proportions_dt)[2:ncol(doc_topic_proportions_dt)] <- paste0('Topic',colnames(doc_topic_proportions_dt)[2:ncol(doc_topic_proportions_dt)]) # Assign topics as column names

topic2doc <- doc_topic_proportions_dt[order(Topic2,decreasing = T)][1:3,]
topic3doc <- doc_topic_proportions_dt[order(Topic3,decreasing = T)][1:3,]

cat("Top 3 documents for topic 2 are:", topic2doc$doc_id)

cat("Top 3 documents for topic 3 are:", topic3doc$doc_id)

# Select rows from original data...
# fb_congress_data[doc_id==1744,]$message
# fb_congress_data[doc_id==4108,]$message

```

Overall, the new model (with $K=3$) does not capture the topics very well.

I prefer $K=50$ over $K=3$ because having only three topics is too coarse and cannot capture the variety and nuances of the different topics discussed.

### 7.

```{r}
posterior_results <- posterior(mylda)

theta <- posterior_results$topics

topic_prevalence <- colMeans(theta)


cat("The most prevalence is topic", which.max(topic_prevalence))

labeled_topics <- c(1:length(topic_prevalence))

plot_data <- data.frame(
  topic = paste0("Topic ", labeled_topics),
  prevalence = topic_prevalence[labeled_topics]
)

ggplot(plot_data, aes(x = reorder(topic, prevalence), y = prevalence)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(
    x = "Topic",
    y = "Avg prevelence",
    title = "Topic and avg prevelence"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 7))

plot_data_subset <- plot_data[c(3, 4, 16, 30, 36), ] |> 
  mutate(
    topic = c(3, 4, 16, 30, 36),
    topic_label = factor(topic, levels = c(3, 4, 16, 30, 36), labels = topic_name))

ggplot(plot_data_subset, aes(x = reorder(topic_label, prevalence), y = prevalence)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(
    x = "Topic",
    y = "Avg prevelence",
    title = "Topic and avg prevelence"
  ) +
  theme_minimal()
```

The most common topic is topic 33.

```{r}
party_raw <- docvars(dtm, "party")

K <- ncol(theta)
colnames(theta) <- paste0("Topic ", 1:K)

theta_df <- as.data.frame(theta) %>%
  mutate(party = party_raw)

theta_long <- theta_df %>%
  pivot_longer(cols = starts_with("Topic"),
               names_to = "topic",
               values_to = "proportion")

labeled_topics <- paste0("Topic ", c(3, 4, 16, 30, 36))

theta_long_labeled <- theta_long %>% filter(topic %in% labeled_topics)

mapping <- c(
  "Topic 3"  = "national security",
  "Topic 4"  = "drug",
  "Topic 16" = "education",
  "Topic 30" = "agriculture",
  "Topic 36" = "healthcare system"
)

theta_long_labeled$topic_label <- mapping[theta_long_labeled$topic]

ggplot(theta_long_labeled,
       aes(x = topic_label, y = proportion, fill = party)) +
  stat_summary(fun = mean, geom = "bar", position = "dodge") +
  labs(x = "Topic", y = "Mean proportion",
       title = "Democrats vs Republicans: Topic prevalence") +
  theme_minimal() +
  coord_flip()
```
From the figure above, we can see that there are clear differences in the prevalence of national security, healthcare system, and agriculture topics between Democrats and Republicans.

```{r}
# glm

glm_topic30 <- glm(`Topic 30` ~ party, family = "quasibinomial", data = theta_df)
# summary(glm_topic30)$coefficients
# broom::tidy(glm_topic30, exponentiate = TRUE)

glm_results <- lapply(labeled_topics, function(tnm) {
  frm <- as.formula(paste0("`", tnm, "` ~ party"))
  fit <- glm(frm, family = "quasibinomial", data = theta_df)
  broom::tidy(fit, exponentiate = TRUE) |> mutate(topic = tnm)
}) |> bind_rows()

glm_wide <- glm_results |> 
  select(topic, term, estimate) |> 
  pivot_wider(names_from = term, values_from = estimate) |> 
  mutate(topic_label = mapping[topic]) |> 
  relocate(topic_label, .after = topic)

glm_wide |> kable()
```

For topic 3, the odds ratio is 1.135, indicating that Republican members are about 13.5% more likely to post about this topic than Democratic members.
In contrast, for topic 36, the odds ratio is 0.850, meaning that Republican members are about 15% less likely to engage with this topic compared to Democrats.
The odds ratios for topics 4, 16, and 30 are closer to 1, suggesting that the party differences for these topics are relatively small.

```{r}
# t test
res <- t.test(theta_df[["Topic 3"]] ~ theta_df$party)
# res$p.value

t_results <- lapply(labeled_topics, function(tnm){
  res <- t.test(theta_df[[tnm]] ~ theta_df$party)
  data.frame(
    topic = tnm,
    p_value = res$p.value,
    mean_in_Democrat = res$estimate[1],
    mean_in_Republican = res$estimate[2]
  )
}
  ) |> 
  bind_rows()
rownames(t_results) <- NULL

t_results <- t_results |> 
  mutate(topic_label = mapping[topic]) |> 
  relocate(topic_label, .after = topic)

t_results |> kable()
```
According to the results of the t-test, topics 3 and 30 are issues that Republicans care more about, while topic 36 is an issue that Democrats care more about. In contrast, there are no significant differences for topics 4 and 16.


### 8.

```{r}
set.seed(1984)
row_ids <- 1:nrow(dtm)
train_ids <- sample(x = row_ids, size = 0.8*length(row_ids), replace = F)
test_ids <- row_ids[!row_ids %in% train_ids]
train_dt <- dtm[train_ids,]
test_dt <- dtm[-train_ids,]

ks <- c(3, 10, 25, 50, 75, 100, 200)
perp <- c()
for(i in 1:length(ks)){
  current_tm <- LDA(x = train_dt,
                    k = ks[i], 
                    method="Gibbs",
                    control=list(iter = 500, 
                                 seed = 1, 
                                 verbose = 0))
  # Compute perplexity
  perp[i] <- perplexity(object = current_tm, 
                        newdata = as.simple_triplet_matrix(test_dt)) # For some reason, perplexity() wants the test data in simple_triplet_matrix format.
  print(perp)
}
plot(x=ks,y=perp)
```

According to the elbow method, $K = 50$ is the best choice because it is the last point where perplexity drops significantly. Increasing K beyond this point does not lead to a substantial decrease in perplexity. In practice, however, one should also consider which value is more interpretable before making the final choice.

## Part 2:

### 1.
```{r}
fb_congress_data <- fread("~/Documents/ML/Labs/Lab5/fb-congress-data3.csv")
```

```{r}
posts_corpus <- corpus(x = fb_congress_data,
                       text_field = "message",
                       meta = list("screen_name", "party"),
                       docid_field = 'doc_id')

# print(posts_corpus)

# Tokenize & clean from particular types of words
mytokens <- tokens(x = posts_corpus, 
                   remove_punct = TRUE, 
                   remove_numbers = TRUE, 
                   remove_symbols = TRUE,
                   remove_url = TRUE,
                   padding = FALSE)
# print(mytokens)

# Make tokens lowercase
mytokens <- tokens_tolower(x = mytokens)

# How does it look now?
# mytokens[1:3]

mysentences <- sapply(mytokens, function(x) paste(x, collapse = " "))
# mysentences[1:3]
```

### 2.
```{r}
set.seed(1984)
w2v <- word2vec(x = mysentences,          # Your data 
                            type = "cbow",    # Model type (the one we talked about in lecture)
                            window = 15,       # Context defined in terms of +-5 words.
                            dim = 50,         # Dimensionality of embeddings
                            iter = 50,        # Estimation iterations (higher means more time...)
                            hs = FALSE,       # Setting to FALSE here --> "negative sampling procedure" (learn difference between fake/true sentences)
                            negative = 15)   # Number of negative samples (how many fake sentences per true sentence)

# w2v_association <- word2vec(x = mysentences,          # Your data 
#                             type =  "skip-gram",    # Model type (the one we talked about in lecture)
#                             window = 15,       # Context defined in terms of +-5 words.
#                             dim = 50,         # Dimensionality of embeddings
#                             iter = 50,        # Estimation iterations (higher means more time...)
#                             hs = FALSE,       # Setting to FALSE here --> "negative sampling procedure" (learn difference between fake/true sentences)
#                             negative = 15)   # Number of negative samples (how many fake sentences per true sentence)
```

### 3.
```{r}
selected_words <- c("russia", "tax", "trump")
w2v_predict <- predict(w2v, selected_words, type = "nearest", top_n = 10)

w2v_predict
```
The prediction results are relatively consistent with expectations. However, in fact, a model trained using the skip-gram approach would produce results that align even more closely with expectations.

### 4.
```{r}
embedding <- as.matrix(w2v)

embedding[rownames(embedding)=="women",]

kingtowoman <- embedding["king", ] - embedding["man", ] + embedding["woman", ]
sims <- word2vec_similarity(embedding, matrix(kingtowoman, nrow = 1))

head(sort(sims[, 1], decreasing = TRUE), 20)

pretrained <- readRDS("~/Documents/ML/Labs/Lab5/glove6B200d.rds")
dim(pretrained)

# embedding_pretained <- as.matrix(pretrained) # pretrained is "matrix" "array" 
# pretrained[rownames(pretrained)=="king",]
kingtowoman <- pretrained["king", ] - pretrained["man", ] + pretrained["woman", ]
sims <- word2vec_similarity(pretrained, matrix(kingtowoman, nrow = 1))
head(sort(sims[, 1], decreasing = TRUE), 20)
```

```{r}
dist_vec <- function(a,b) sqrt(sum((a-b)^2))
dist_sta_man <- dist_vec(pretrained["statistician", ], pretrained["man", ])
dist_sta_woman <- dist_vec(pretrained["statistician", ], pretrained["woman", ])
occupationalbias <- dist_sta_man - dist_sta_woman
occupationalbias
```

The occupational bias is `r round(occupationalbias, 3)`, indicating that "statistician" is farther from "woman" than from "man" in the embedding space, which suggests the presence of occupational bias.


### 5.
```{r}
corpus_dem <- corpus_subset(posts_corpus, party == "Democrat")
corpus_rep <- corpus_subset(posts_corpus, party == "Republican")

mytokens_dem <- tokens(x = corpus_dem, 
                   remove_punct = TRUE, 
                   remove_numbers = TRUE, 
                   remove_symbols = TRUE,
                   remove_url = TRUE,
                   padding = FALSE)

mytokens_rep <- tokens(x = corpus_rep, 
                   remove_punct = TRUE, 
                   remove_numbers = TRUE, 
                   remove_symbols = TRUE,
                   remove_url = TRUE,
                   padding = FALSE)

# Make tokens lowercase
mytokens_dem <- tokens_tolower(x = mytokens_dem)
mytokens_rep <- tokens_tolower(x = mytokens_rep)

mysentences_dem <- sapply(mytokens_dem, function(x) paste(x, collapse = " "))
mysentences_rep <- sapply(mytokens_rep, function(x) paste(x, collapse = " "))
```

```{r}
set.seed(1984)
w2v_dem <- word2vec(x = mysentences_dem,          # Your data 
                            type = "cbow",    # Model type (the one we talked about in lecture)
                            window = 15,       # Context defined in terms of +-5 words.
                            dim = 50,         # Dimensionality of embeddings
                            iter = 50,        # Estimation iterations (higher means more time...)
                            hs = FALSE,       # Setting to FALSE here --> "negative sampling procedure" (learn difference between fake/true sentences)
                            negative = 15)   # Number of negative samples (how many fake sentences per true sentence)

w2v_rep <- word2vec(x = mysentences_rep,          # Your data 
                            type = "cbow",    # Model type (the one we talked about in lecture)
                            window = 15,       # Context defined in terms of +-5 words.
                            dim = 50,         # Dimensionality of embeddings
                            iter = 50,        # Estimation iterations (higher means more time...)
                            hs = FALSE,       # Setting to FALSE here --> "negative sampling procedure" (learn difference between fake/true sentences)
                            negative = 15)   # Number of negative samples (how many fake sentences per true sentence)
```

```{r}
predict(w2v_dem, c("immigrants", "abortion"), type = "nearest", top_n = 10)

predict(w2v_rep, c("immigrants", "abortion"), type = "nearest", top_n = 10)
```
The results are consistent with expectations. For Republicans, the nearest words to "immigrants" are more negative, while for Democrats, the nearest words are relatively more neutral or positive. The same pattern is observed for "abortion", which reflects the highly conservative beliefs of the Republican Party.

### 6
```{r}
positive <- fread("~/Documents/ML/Labs/Lab5/positive.txt")[[1]]
negative <- fread("~/Documents/ML/Labs/Lab5/negative.txt")[[1]]
selected_words <- c("russia", "tax", "trump")
```

```{r}
embeddings_dem <- as.matrix(w2v_dem)
pos_in_dem <- intersect(positive, rownames(embeddings_dem))
neg_in_dem <- intersect(negative, rownames(embeddings_dem))
pos_matrix_dem <- embeddings_dem[pos_in_dem, , drop = FALSE]
neg_matrix_dem <- embeddings_dem[neg_in_dem, , drop = FALSE]
pos_vec_dem <- colMeans(pos_matrix_dem)
neg_vec_dem <- colMeans(neg_matrix_dem)

sentiment_vec_dem <- neg_vec_dem - pos_vec_dem
sims_dem <- word2vec_similarity(embeddings_dem, sentiment_vec_dem)
similarity_table_dem <- data.frame(
  word = rownames(embeddings_dem),
  sim = as.numeric(sims_dem[, 1])
)
sims_dem_filtered <- similarity_table_dem |> 
  filter(!(word %in% negative)) |> 
  arrange(desc(sim)) |> 
  mutate(rank = row_number())
```


```{r}
embeddings_rep <- as.matrix(w2v_rep)
pos_in_rep <- intersect(positive, rownames(embeddings_rep))
neg_in_rep <- intersect(negative, rownames(embeddings_rep))
pos_matrix_rep <- embeddings_rep[pos_in_rep, , drop = FALSE]
neg_matrix_rep <- embeddings_rep[neg_in_rep, , drop = FALSE]
pos_vec_rep <- colMeans(pos_matrix_rep)
neg_vec_rep <- colMeans(neg_matrix_rep)

sentiment_vec_rep <- neg_vec_rep - pos_vec_rep
sims_rep <- word2vec_similarity(embeddings_rep, sentiment_vec_rep)
similarity_table_rep <- data.frame(
  word = rownames(embeddings_rep),
  sim = as.numeric(sims_rep[, 1])
)
sims_rep_filtered <- similarity_table_rep |> 
  filter(!(word %in% negative)) |> 
  arrange(desc(sim)) |> 
  mutate(rank = row_number())
```

```{r}
sims_dem_filtered |> head(10)
sims_rep_filtered |> head(10)

rank_dem <- sims_dem_filtered |> filter(word %in% selected_words)

rank_rep <- sims_rep_filtered |> filter(word %in% selected_words)

rank_dem |> 
  kable(caption = "Word Rank for Democrats")
rank_rep |> 
  kable(caption = "Word Rank for Republicans")
```

I selected trump, russia, gun, and tax. For example, we can observe that trump has a lower similarity score on the negative sentiment dimension for Republicans, meaning that trump is relatively less negative. In contrast, trump has a higher similarity score on the negative sentiment dimension for Democrats, indicating that trump is perceived as highly negative.

### 7
```{r}
fit_and_rank <- function(mysentences, positive_words = postive, negative_words = negative, type = "cbow", window = 15, dim = 50, iter = 50, hs = FALSE, negative = 15){
  w2v <- word2vec(x = mysentences, type = type, window = window, dim = dim,
                            iter = iter, hs = hs, negative = negative)
  emb <- as.matrix(w2v)
  pos_in <- intersect(positive_words, rownames(emb))
  neg_in <- intersect(negative_words, rownames(emb))
  pos_vec <- colMeans(emb[pos_in, , drop = FALSE])
  neg_vec <- colMeans(emb[neg_in, , drop = FALSE])
  sentiment_vec <- neg_vec - pos_vec
  sims <- word2vec_similarity(emb, sentiment_vec)
  similarity_table <- data.frame(
  word = rownames(emb),
  sim = as.numeric(sims[, 1])
)
sims_filtered <- similarity_table |> 
  filter(!(word %in% negative_words)) |> 
  arrange(desc(sim)) |> 
  mutate(rank = row_number())
}

bootstrap_by_party <- function(mysentences, positive_words = positive, negative_words = negative, n_boot = 20, seed = 1984){
  set.seed(seed)
  map_dfr(seq_len(n_boot), function(bi) {
    idx <- sample.int(length(mysentences), replace = TRUE)
    sent_boot <- mysentences[idx]
    fit_and_rank(sent_boot, positive_words = positive, negative_words = negative) |>
      filter(word %in% selected_words) |>
      mutate(iter = bi)
  })
}
```

```{r}
boot_dem <- bootstrap_by_party(mysentences_dem) |>
  mutate(party = "Democrat")
boot_rep <- bootstrap_by_party(mysentences_rep) |>
  mutate(party = "Republican")

boot_all <- bind_rows(boot_dem, boot_rep)

# saveRDS(boot_all, file = "boot_all.rds")


boot_summary <- boot_all |>
  group_by(party, word) |>
  summarize(n_iter = n(),
            sim_mean = mean(sim, na.rm = TRUE),
            sim_low = quantile(sim, 0.025, na.rm = TRUE),
            sim_high = quantile(sim, 0.975, na.rm = TRUE),
            rank_mean = mean(rank, na.rm = TRUE),
            .groups = "drop")

boot_sim_wide <- boot_summary |>
  select(party, word, sim_mean, sim_low, sim_high) |>
  pivot_wider(names_from = party, values_from = c(sim_mean, sim_low, sim_high))

boot_summary |> kable()
boot_sim_wide |> kable()
```
The conclusion is consistent with the above, but we can observe that only russia has a lower bound greater than zero. This indicates that russia is the only word with a significant difference between the two parties, while the differences for the other words are not statistically significant.
