---
title: "Machine Learning Lab"
author: "Yukun Jiao"
date: "2025-08-20"
toc: false
format: pdf
editor: 
  markdown: 
    wrap: 72
---

# Lab 1

## Part 1: Bernie Sanders and Donald Trump tweets.

### 1.

```{r}
#| warning: false
#| message: false

library(data.table)
library(glmnet)
library(ggplot2)
library(mlbench)
library(caret)
library(splines)
library(ggeffects)
```

```{r}
# load data
trumpbernie <- fread(file = "~/Documents/ML/Labs/Lab1/trumpbernie.csv")
dim(trumpbernie) 
```

The dataset consists of 1003 observations and 1496 variables. Since the number of variables substantially exceeds the number of observations, the dataset is high-dimensional. Under such conditions, a standard logistic regression model is likely to perform poorly for prediction, due to overfitting and multicollinearity.

### 2.

```{r}
mymodel <- glm(trump_tweet ~ ., data = trumpbernie, family="binomial")
sum(!is.na(coef(mymodel)))
coef(mymodel)[(1001:1050)]
```

There are many coefficients that are missing. In fact, 495 variables were identified as singular, and their coefficients could not be estimated due to perfect collinearity with other predictors.

```{r}
# Extract predictions on training data & observed values
comparison_df <- data.frame(train_predictions=mymodel$fitted.values,
observed=mymodel$y)
# Apply prediction threshold
comparison_df$train_predictions<-ifelse(comparison_df$train_predictions>=0.5,
yes = 1,
no = 0)
# Compute accuracy (scale: 0-1, 0=0%, 1=100%)
nrow(comparison_df[comparison_df$train_predictions==comparison_df$observed,]) /
nrow(comparison_df)
```

The training accuracy here is 1 (100%). This indicates that the model is overfitted, and in such cases, the predictive capacity of the model is often poor.

### 3.

```{r}
#| warning: false
#| message: false


# format the outcome variable as a factor variable
trumpbernie$trump_tweet <- factor(trumpbernie$trump_tweet, levels = c(0,1))

# Use package "caret" to do k-fold cross-validation
tc <- caret::trainControl(method = 'cv', number = 3)

set.seed(12345) # Set to ensure that folds are the same across models
linear_model <- caret::train(trump_tweet ~ .,
                             data = trumpbernie,
                             method = "glm",
                             trControl = tc,
                             family = "binomial")
```

Some warnings were issued during model fitting, including non-convergence of the algorithm and rank-deficiency in the fitted model.

```{r}
linear_model$results$Accuracy

pred_class <- predict(linear_model, newdata = trumpbernie)
train_acc <- mean(pred_class == trumpbernie$trump_tweet)
train_acc
```

The model achieved a test accuracy of `r linear_model$results$Accuracy` indicating that its predictions are essentially no better than random guessing, which is consistent with my expectations. These results suggest that the model is severely overfitted.

### 4.
```{r}
X <- trumpbernie[,-c('trump_tweet'),with=F] # Exclude response and id columns.
X <- as.matrix(X) # Make input data into a matrix

cvglmnet <- cv.glmnet(x = X, 
                            y = trumpbernie$trump_tweet,
                            nfolds = 5,
                            standardize = TRUE,
                            family='binomial',
                            alpha=0,
                            type.measure = 'class')

cvglmnet$lambda.min
cvglmnet$cvm[cvglmnet$index[[1]]]
```

The value of $\lambda$ that minimizes the test error is `r cvglmnet$lambda.min`, corresponding to a test accuracy of `r 1 - cvglmnet$cvm[cvglmnet$index[[1]]]`, indicating that it is a better-performing model. The previous model had higher variance because its greater flexibility caused the results to be highly sensitive, leading to larger variance.


### 5.
```{r}
plot(cvglmnet, sign.lambda = 1)
```

As $\lambda$ increases, the model becomes less flexible, and the variance decreases while the bias increases. Generally, the reduction in variance outweighs the increase in bias until the value of $\lambda$ that minimizes the test error is reached, which represents a noteworthy trade-off.


### 6.
```{r}
best_coefs <- coef(cvglmnet, s = "lambda.min")
best_coefs_dt <- data.table(word=rownames(best_coefs),
                            coef=best_coefs[,1])
best_coefs_dt[order(coef,decreasing = T)]
```

After extracting the coefficients at lambda.min, I found that words with positive coefficients (such as atlanta, patriot, and colorado) are more likely to indicate a Trump tweet, while words with negative coefficients (such as vulner, volunt, and view) are more likely to indicate a Bernie tweet. This is generally consistent with my expectations regarding the language styles of the two politicians: Trump's tweets tend to focus on geographic locations or patriotic topics, while Bernie's tweets highlight progressive social themes, emphasizing social vulnerability and volunteer-related activities.


## Part 2: Social Network Ad Purchase

### 1.

```{r}
sna_data <- fread("~/Documents/ML/Labs/Lab1/Kaggle_Social_Network_Ads.csv")
sna_data$Purchased <- factor(sna_data$Purchased, levels = c(0,1))
```

### 2.

```{r}
# Use package "caret" to do k-fold cross-validation
tc <- caret::trainControl(method = 'cv', number = 5)

set.seed(12345) # Set to ensure that folds are the same across models
nonlinear_model <- caret::train(Purchased ~ .,
                             data = sna_data,
                             method = "glm",
                             trControl = tc,
                             family = "binomial")

nonlinear_model$results
```
The test accuracy is `r nonlinear_model$results$Accuracy`.

### 3.

```{r}
set.seed(12345) # Set to ensure that folds are the same across models
gam_model1 <- caret::train(Purchased ~ ns(Age,2) + ns(Salary, 2) + Gender,
                           data = sna_data,
                           method = "glm",
                           trControl = tc,
                           family = "binomial")
gam_model1$results

gam_model2 <- caret::train(Purchased ~ ns(Age,3) + ns(Salary, 3) + Gender,
                           data = sna_data,
                           method = "glm",
                           trControl = tc,
                           family = "binomial")
gam_model2$results

gam_model3 <- caret::train(Purchased ~ ns(Age,4) + ns(Salary, 4) + Gender,
                           data = sna_data,
                           method = "glm",
                           trControl = tc,
                           family = "binomial")
gam_model3$results
```

#### a.
The test accuracy has improved, confirming the effectiveness of the updated model.

#### b.
This indicates that the previous standard logistic regression model was underfitting. In other words, the earlier linear model was insufficient to capture the patterns in this dataset, so we need a more flexible model to fit it---namely, the GAM. Compared to the linear model, the GAM introduces more bias.

#### c.
I would select GAM 3, as it achieves the highest test accuracy among these models.

### 4.
```{r}
final_model <- glm(Purchased ~ ns(Age,4) + ns(Salary, 4) +  Gender, 
                   data=sna_data, 
                   family = "binomial")
ggpreds <- ggpredict(final_model)
plot(ggpreds)
```

The results show that the effects of Age and Salary on the probability of purchase are both nonlinear, indicating that a standard linear logistic regression cannot fully capture these nonlinear effects.

### 5.

No, because Ridge and Lasso are intended to increase bias in order to mitigate overfitting. However, the current situation involves underfitting, which requires a reduction in bias.

## Part 3: K-fold cross validation from scratch (bonus!)

```{r}

dt <- readRDS("~/Documents/ML/Labs/Lab1/dt.rds")

head(dt)
```

```{r}
chunk2 <- function(x,n) split(x, cut(seq_along(x), n, labels = FALSE))

cv_poly <- function(k, d, dt) {
  n <- nrow(dt)
  folds <- chunk2(1:n, k)
  mse_list <- c()
  for (i in 1:k){
    test_idx <- folds[[i]]
    train_idx <- setdiff(1:n, test_idx)
    train_data <- dt[train_idx, ]
    test_data  <- dt[test_idx, ]
    
     model <- lm(Y ~ poly(X, d), data = train_data)
     preds <- predict(model, newdata = test_data)
     
     mse <- mean((test_data$Y - preds)^2)
     mse_list <- c(mse_list, mse)
  }
  return(mean(mse_list))
}

d_values <- 1:10
mse_values <- sapply(d_values, function(d) cv_poly(5, d, dt))
plot(d_values, mse_values, type = "b",
     pch = 19, col = "darkred", lty = 1,
     xlab = "Polynomial Degree (d)",
     ylab = "Average MSE",
     main = "Average MSE Across Polynomial Degrees",
     xaxt = "n")

axis(1, at = d_values, labels = d_values)
```

The results show that as the polynomial degree $d$ increases, the average \text{MSE} initially decreases and then rises, indicating that the model captures some nonlinearity in the data, but overly high-degree polynomials may lead to overfitting. The minimum average \text{MSE} is observed at $d = 10$; however, this may reflect overfitting due to the small dataset. Choosing lower-degree polynomials such as $d = 3$ or $d = 5$ achieves comparably low average \text{MSE} while ensuring a more robust and interpretable model.
