---
title: "Machine Learning Lab"
author: "Yukun Jiao"
date: "2025-09-03"
toc: false
format: pdf
knitr: 
  opts_chunk:
    fig.align: center
editor: 
  markdown: 
    wrap: 72
---

# Lab 3

## Part 1:

### 1.

```{r}
#| warning: false
#| message: false

library(keras); library(keras3); library(tensorflow)
library(ggplot2); library(dplyr); library(tidyr)
library(data.table); library(rpart); library(caret)
```


```{r}
adults <- readRDS("~/Documents/ML/Labs/Lab3/adults.rds")
adults <- na.omit(adults)
# X_all <- scale(adults[, -which(colnames(adults) == "y")])
X_all <- adults[, -which(colnames(adults) == "y")]
# X_all <- scale(adults[, -6])
y_all <- adults[, "y"]       # labels (0/1)

set.seed(42)
n <- nrow(X_all)
idx <- sample(seq_len(n), size = floor(0.8 * n))
X_train <- X_all[idx, , drop = FALSE]
y_train <- y_all[idx]
X_test  <- X_all[-idx, , drop = FALSE]
y_test  <- y_all[-idx]
```

Based on the information that there are no complex interactions or non-linearities, it can be inferred that a neural network will not excel on this dataset. Neural networks are better suited for capturing non-linear relationships and interactions. Although they can also capture linear relationships, they require more parameters, lack interpretability, and their predictive performance is likely to be close to that of a simple linear regression model.

### 2.

```{r}
#| warning: false
#| message: false

set.seed(1984)
train_df <- data.frame(y = y_train, X_train)
test_df <- data.frame(y = y_test, X_test)

salary_glm <- glm(y ~ ., data = train_df, family = "binomial")

p_hat <- predict(salary_glm, newdata = test_df, type = "response")


acc_bin <- function(p, y_true, thr = 0.5) mean((p > thr) == y_true)
acc_bin(p_hat, test_df$y, thr = 0.5)

# y_pred <- ifelse(p_hat >= 0.5, 1, 0)
# accuracy <- mean(y_pred == test_df$y)
# accuracy
```

The accuracy is `r acc_bin(p_hat, test_df$y, thr = 0.5)`.

### 3.

```{r}
input_dim <- ncol(X_train)

k_clear_session()
model_A <- keras_model_sequential() |>
  # Hidden layer:
  # units=32 (width of layer), activation="relu" (nonlinearity)
  layer_dense(units = 5, activation = "relu", input_shape = input_dim) |>
  # Output layer for binary classification:
  layer_dense(units = 1, activation = "sigmoid")

model_A |>
  compile(
    optimizer = optimizer_adam(1e-3),   # optimizer/learning rate
    loss = "binary_crossentropy",
    metrics = "accuracy"
  )

history_A <- model_A |>
  fit(
    X_train, y_train,
    epochs = 25,
    batch_size = 32,
    validation_split = 0.2,
    verbose = 0
  )

proba_A <- as.numeric(model_A |> predict(X_test, verbose = 0))
acc_A   <- acc_bin(proba_A, y_test)
cat(sprintf("[adults] Model 1 (1 hidden) — test accuracy: %.3f\n", acc_A))

```

The accuracy of the MNN is slightly better compared to a standard logistic regression model, but still very close. Moreover, it requires more computational power and parameter tuning, and it lacks interpretability. This is consistent with my assessment in #1.

### 4.

```{r}
#| warning: false
#| message: false

k_clear_session()
model_D <- keras_model_sequential() |>
  layer_dense(units = 256, activation = "relu", input_shape = input_dim) |>
  layer_dense(units = 128, activation = "relu") |>
  layer_dense(units = 64, activation = "relu") |> 
  layer_dense(units = 32, activation = "relu") |>
  layer_dense(units = 1, activation = "sigmoid")

model_D |>
  compile(
    optimizer = optimizer_adam(1e-3),
    loss = "binary_crossentropy",
    metrics = "accuracy"
  )

history_D <- model_D |>
  fit(
    X_train, y_train,
    epochs = 25,
    batch_size = 32,
    validation_split = 0.2,
    verbose = 0
  )

proba_D <- as.numeric(model_D |> predict(X_test, verbose = 0))
acc_D   <- acc_bin(proba_D, y_test)
cat(sprintf("[adults] Model 2 (4 hidden) — test accuracy: %.3f\n", acc_D))
# model_D |> summary()
```
The number of trainable parameters is $1536 + 32896 + 8256 + 2080 + 33 = 44,801$.

The performance is better, but the improvement is not very large. I suspect there may be some limited non-linear relationships and interactions that allow the neural network to capture them, leading to the improvement in accuracy.

### 5.

```{r}
adults_aug <- readRDS("~/Documents/ML/Labs/Lab3/adults_aug.rds")
adults_aug <- na.omit(adults_aug)
X_all <- adults_aug[, -which(colnames(adults_aug) == "y")]
y_all <- adults_aug[, "y"]       # labels (0/1)

set.seed(42)
n <- nrow(X_all)
idx <- sample(seq_len(n), size = floor(0.8 * n))
X_train <- X_all[idx, , drop = FALSE]
y_train <- y_all[idx]
X_test  <- X_all[-idx, , drop = FALSE]
y_test  <- y_all[-idx]
```

```{r}
#| warning: false
#| message: false

set.seed(1984)
train_df <- data.frame(y = y_train, X_train)
test_df <- data.frame(y = y_test, X_test)

salary_glm <- glm(y ~ ., data = train_df, family = "binomial")

p_hat <- predict(salary_glm, newdata = test_df, type = "response")


acc_bin <- function(p, y_true, thr = 0.5) mean((p > thr) == y_true)
acc_bin(p_hat, test_df$y, thr = 0.5)
```

```{r}
input_dim <- ncol(X_train)

k_clear_session()
model_A <- keras_model_sequential() |>
  # Hidden layer:
  # units=32 (width of layer), activation="relu" (nonlinearity)
  layer_dense(units = 5, activation = "relu", input_shape = input_dim) |>
  # Output layer for binary classification:
  layer_dense(units = 1, activation = "sigmoid")

model_A |>
  compile(
    optimizer = optimizer_adam(1e-3),   # optimizer/learning rate
    loss = "binary_crossentropy",
    metrics = "accuracy"
  )

history_A <- model_A |>
  fit(
    X_train, y_train,
    epochs = 25,
    batch_size = 32,
    validation_split = 0.2,
    verbose = 0
  )

proba_A <- as.numeric(model_A |> predict(X_test, verbose = 0))
acc_A   <- acc_bin(proba_A, y_test)
cat(sprintf("[adults_aug] Model 3 (1 hidden) — test accuracy: %.3f\n", acc_A))
```

```{r}
#| warning: false
#| message: false

k_clear_session()
model_D <- keras_model_sequential() |>
  layer_dense(units = 256, activation = "relu", input_shape = input_dim) |>
  layer_dense(units = 128, activation = "relu") |>
  layer_dense(units = 64, activation = "relu") |> 
  layer_dense(units = 32, activation = "relu") |>
  layer_dense(units = 1, activation = "sigmoid")

model_D |>
  compile(
    optimizer = optimizer_adam(1e-3),
    loss = "binary_crossentropy",
    metrics = "accuracy"
  )

history_D <- model_D |>
  fit(
    X_train, y_train,
    epochs = 25,
    batch_size = 32,
    validation_split = 0.2,
    verbose = 0
  )

proba_D <- as.numeric(model_D |> predict(X_test, verbose = 0))
acc_D   <- acc_bin(proba_D, y_test)
cat(sprintf("[adults_aug] Model 4 (4 hidden) — test accuracy: %.3f\n", acc_D))
# model_D |> summary()
```

The neural network model should significantly outperform the standard logistic regression model.

On this augmented dataset, the neural network model substantially improves accuracy. This is because many variables containing non-linear relationships and interactions were added to the dataset. Neural networks can capture these effects very well, whereas the standard logistic regression model cannot, and therefore its accuracy is lower than that of the neural network model.

### 6.


```{r}
set.seed(1984)

k_clear_session()
model_E <- keras_model_sequential() |>
  layer_dense(units = 256, activation = "relu", input_shape = input_dim) |>
  layer_dropout(rate = 0.1) |>
  layer_dense(units = 128, activation = "relu") |>
  layer_dropout(rate = 0.1) |>
  layer_dense(units = 64, activation = "relu") |> 
  layer_dropout(rate = 0.1) |>
  layer_dense(units = 32, activation = "relu") |>
  layer_dense(units = 1, activation = "sigmoid")

model_E |>
  compile(
    optimizer = optimizer_adam(1e-3),
    loss = "binary_crossentropy",
    metrics = "accuracy"
  )

history_E <- model_E |>
  fit(
    X_train, y_train,
    epochs = 25,
    batch_size = 32,
    validation_split = 0.2,
    verbose = 0
  )

proba_E <- as.numeric(model_E |> predict(X_test, verbose = 0))
acc_E   <- acc_bin(proba_E, y_test)
cat(sprintf("[adults_aug] Model 5 (4 hidden) — test accuracy: %.3f\n", acc_E))
# model_E |> summary()

# results <- model_E |> evaluate(X_test, y_test, verbose = 0)
# results$accuracy
```


```{r}
k_clear_session()
model_F <- keras_model_sequential() |>
  layer_dense(units = 256, activation = "relu", input_shape = input_dim) |>
  layer_dropout(rate = 0.9) |>
  layer_dense(units = 128, activation = "relu") |>
  layer_dropout(rate = 0.9) |>
  layer_dense(units = 64, activation = "relu") |> 
  layer_dropout(rate = 0.9) |>
  layer_dense(units = 32, activation = "relu") |>
  layer_dense(units = 1, activation = "sigmoid")

model_F |>
  compile(
    optimizer = optimizer_adam(1e-3),
    loss = "binary_crossentropy",
    metrics = "accuracy"
  )

history_F <- model_F |>
  fit(
    X_train, y_train,
    epochs = 25,
    batch_size = 32,
    validation_split = 0.2,
    verbose = 0
  )

proba_F <- as.numeric(model_F |> predict(X_test, verbose = 0))
acc_F   <- acc_bin(proba_F, y_test)
cat(sprintf("[adults_aug] Model 6 (4 hidden) — test accuracy: %.3f\n", acc_F))
# model_F |> summary()
```

A dropout rate of 0.1 performs better, possibly because dropout learning provides regularization and reduces variance. Although it may increase bias, the reduction in variance outweighs the increase in bias, which makes it worthwhile, and accuracy is further improved.

With a dropout rate of 0.9, accuracy drops dramatically. This is likely also due to reduced variance but with a much greater increase in bias. The increase in bias clearly outweighs the benefits of reduced variance, leading to a sharp decline in accuracy.


## Part 2

### 1.

```{r}
fashion_train <- readRDS("~/Documents/ML/Labs/Lab3/fashion_2016_train.rds")
fashion_test <- readRDS("~/Documents/ML/Labs/Lab3/fashion_2016_test.rds")
```

### 2.

take a look at these labels
```{r}
x_train <- fashion_train$images
y_train <- fashion_train$labels
x_test <- fashion_test$images
y_test <- fashion_test$labels


# fashion_train$labels
num <- 5
# fashion_label <- fashion_train$labels[num]
# fashion_train$fashion_labels[fashion_label+1]

par(mar = c(0,0,0,0))
test <- x_train[num,,,]
item_name <- fashion_train$fashion_labels[fashion_train$labels[num]+1]
image(t(apply(test, 2, rev)), col = gray.colors(256), axes = FALSE,
      main = item_name)
print(item_name)

```

```{r}
img_rows <- 28L; img_cols <- 28L; channels <- 1L
x_train <- array_reshape(x_train, c(nrow(x_train), img_rows, img_cols, channels)) # Note: only new part here is the channels (in our case, it is gray-scale; so only 1 channel; gets more interesting with color)
x_test  <- array_reshape(x_test,  c(nrow(x_test),  img_rows, img_cols, channels))
# x_train <- x_train / 255 # normalize by dividing by max-value (1=most dark; 0=most bright)
# x_test  <- x_test  / 255
y_train_cat <- to_categorical(y_train, num_classes = 10)  # one-hot targets
y_test_cat  <- to_categorical(y_test,  num_classes = 10)
```


```{r}
#| warning: false
#| message: false

k_clear_session()
cnn <- keras_model_sequential() |>
  layer_conv_2d(filters = 8, kernel_size = c(3,3), activation = "relu",
                input_shape = c(img_rows, img_cols, channels)) |>
  layer_max_pooling_2d(pool_size = c(2,2)) |>
  layer_flatten() |>
  # layer_dropout(rate = 0.95) |>
  layer_dense(units = 8, activation = "relu") |>
  layer_dense(units = 10, activation = "softmax")
```

```{r}
set.seed(1984)
cnn |>
  compile(
    optimizer = optimizer_adam(),
    loss      = "sparse_categorical_crossentropy",
    metrics   = "accuracy"
  )

history_cnn <- cnn |>
  fit(
    x_train, y_train,
    epochs = 3,
    batch_size = 128, 
    validation_split = 0.1,
    verbose = 0
  )

# Evaluate on test set
cnn_eval <- cnn |>
  evaluate(x_test, y_test, verbose = 0)
cat(sprintf("\n[fashion] CNN — test accuracy: %.3f\n", as.numeric(cnn_eval['accuracy'])))
```

The accuracy is `r round(cnn_eval$accuracy, 3)`.

Increasing M (number of hidden layers), K (number of convolutional layers), the number of filters, or the number of hidden units in the fully connected layer should improve performance, as it allows the model to capture more local features.

### 3.
```{r}
k_clear_session()
cnn_complex <- keras_model_sequential() |>
  layer_conv_2d(filters = 32, kernel_size = c(3,3), activation = "relu",
                input_shape = c(img_rows, img_cols, channels)) |>
  layer_max_pooling_2d(pool_size = c(2,2)) |>
  layer_conv_2d(filters = 64, kernel_size = c(3,3), activation = "relu",
                input_shape = c(img_rows, img_cols, channels)) |>
  layer_max_pooling_2d(pool_size = c(2,2)) |>
  layer_flatten() |>
  # layer_dropout(rate = 0.95) |>
  layer_dense(units = 64, activation = "relu") |>
  layer_dense(units = 32, activation = "relu") |>
  layer_dense(units = 10, activation = "softmax")

cnn_complex |>
  compile(
    optimizer = optimizer_adam(),
    loss      = "sparse_categorical_crossentropy",
    metrics   = "accuracy"
  )

history_cnn_complex <- cnn_complex |>
  fit(
    x_train, y_train,
    epochs = 3,
    batch_size = 128, 
    validation_split = 0.1,
    verbose = 0
  )

# Evaluate on test set
cnn_complex_eval <- cnn_complex |>
  evaluate(x_test, y_test, verbose = 0)
cat(sprintf("\n[fashion] cnn_complex — test accuracy: %.3f\n", as.numeric(cnn_complex_eval['accuracy'])))
```

The accuracy decreased, likely due to overfitting, as no regularization was applied.

From the perspective of the bias-variance trade-off, the model achieved lower bias, but the variance increased too much, which led to the drop in performance.


```{r}
cnn |> summary()
cnn_complex |> summary()
```


### 4.
Because downsampling increases bias and reduces variance. However, if the downsampling is too aggressive, too much detail will be lost, making it difficult for the filters to effectively capture features.

### 5.

```{r}
predictions <- cnn %>% predict(x_test)
predicted_classes <- apply(predictions, 1, which.max)- 1
accuracy_by_class_dt <- data.table(actual=y_test,predicted=predicted_classes)
accuracy_by_class_dt[,correct := fifelse(actual==predicted, yes = 1, no = 0)]
fashion_labels <- c("T-shirt/top", "Trouser", "Pullover", "Dress", "Coat",
"Sandal", "Shirt", "Sneaker", "Bag", "Ankle boot")
fashion_labels_dt <- data.table(label=fashion_labels,number=0:9)
accuracy_by_class_dt <- merge(x=accuracy_by_class_dt,y=fashion_labels_dt,by.x='actual',by.y='number')
accuracy_by_class_dt[,.(accuracy=mean(correct)),by=label][order(accuracy,decreasing = T)]
```
From the table above, we can see that there are significant differences in predictability across categories. For example, Ankle boot, Sandal, and Trouser are the easiest to classify correctly, whereas Shirt is difficult to classify.

This can be considered reasonable.

### 6.

```{r}
fashion_unlabeled <- readRDS("~/Documents/ML/Labs/Lab3/fashion_2015_2016_unlabeled.rds")

x_test <- fashion_unlabeled$images

predictions <- cnn %>% predict(x_test)
predicted_classes <- apply(predictions, 1, which.max)- 1
fashion_unlabeled$labels <- predicted_classes

demo_dt <- as.data.table(fashion_unlabeled$demographics)
demo_dt[, id := .I]

labels_dt <- data.table(id = seq_along(fashion_unlabeled$labels),
                        labels = fashion_unlabeled$labels)

merged_dt <- merge(demo_dt, labels_dt, by = "id")

fashion_labels_dt

fashion_full_dt <- merge(
  x    = merged_dt,
  y    = fashion_labels_dt,
  by.x = "labels",
  by.y = "number",
  all.x = TRUE
)
```

```{r}
fashion_full_dt
```

2016 到 2017 年间消费者行为模式是如何变化的

预计

```{r}
fashion_full_dt |> 
  group_by(label, year) |> 
  summarize(
    age_average = mean(age),
    income_average = mean(income),
    education_years_average = mean(education_years)
  )
```


```{r}
fashion_full_dt$label <- factor(fashion_full_dt$label)
levels(fashion_full_dt$label)
fashion_full_dt$label <- relevel(fashion_full_dt$label, ref = "Bag")

mn_fit <- multinom(label ~ age + year + ses_score, data = fashion_full_dt, trace = FALSE)


summary(mn_fit)
exp(coef(mn_fit))
```

In general, as the ses_score increases, the probability of choosing a certain product rather than a bag decreases. For example, for ankle boots, each one-unit increase in ses_score reduces the relative risk ratio of choosing ankle boots over bags to 66% of its original value.

On the other hand, age and year have little impact on product choice, as their relative risk ratios for choosing a particular product over bags remain largely unchanged.
