---
title: "Inequality and Segregation Lab 1"
author: "Yukun Jiao"
date: "2025-10-27"
toc: false
format:
  pdf
knitr: 
  opts_chunk:
    fig.align: center
editor: 
  markdown: 
    wrap: 72
execute:
  echo: false
  warning: false
  message: false
---

This report investigates the differences between Sweden and Hungary in their Lorenz curves, Gini coefficients, and the variance decompositions of income, then provides a preliminary analysis of the results. The results indicate that Sweden's income distribution is more equal. The variance decomposition further highlights two distinct patterns.

# Part 1
```{r}
library(kableExtra)
library(wid)
library(ineq)
```

```{r}
# wid_data <- download_wid(
#   indicators = "aptinc",
#   areas = "SE",
#   years = "2022",  
#   perc = "all",
#   ages = "all",
#   pop = "all",
#   metadata = FALSE,
#   include_extrapolations = TRUE,
#   verbose = TRUE
# )

# saveRDS(wid_data, file = "wid_data.RDS")
wid_data <- readRDS("wid_data.RDS")

pattern <- "^p(\\d{1,2})p(\\d{1,2})$"

# Use grepl with an additional condition to check that the difference between the two numbers is exactly 1
filtered_df <- wid_data[grepl(pattern, wid_data$percentile) & 
                          as.numeric(sub("p(\\d{1,2})p(\\d{1,2})", "\\2", wid_data$percentile)) -
                          as.numeric(sub("p(\\d{1,2})p(\\d{1,2})", "\\1", wid_data$percentile)) == 1, ]


#### Calculate Gini:

# Step 1: Sort by income and create the rank (R) variable
filtered_df <- filtered_df[order(filtered_df$value), ]
filtered_df$rank <- seq_along(filtered_df$value) / nrow(filtered_df)

# Step 2: Calculate the mean of the income values (x̄)
mean_income <- mean(filtered_df$value)

# Step 3: Calculate the covariance between income and rank
covar <- cov(filtered_df$value, filtered_df$rank)

# Step 4: Calculate the Gini coefficient using the formula
gini_coefficient_formula <- 2/mean_income * covar
# 
# # Print the Gini coefficient
# print(gini_coefficient)

# Calculate the Gini coefficient using the ineq::Gini function
gini_coefficient <- ineq::Gini(filtered_df$value)

# Display the Gini coefficient
# print(gini_coefficient)
```


```{r}
# wid_data_cn <-download_wid(
#   indicators = "aptinc",
#   areas = "HU",
#   years = "2022",
#   perc = "all",
#   ages = "all",
#   pop = "all",
#   metadata = FALSE,
#   include_extrapolations = TRUE,
#   verbose = TRUE
# )

# saveRDS(wid_data_cn, file = "wid_data_cn.RDS")
wid_data_cn <- readRDS("wid_data_cn.RDS")

pattern <- "^p(\\d{1,2})p(\\d{1,2})$"

# Use grepl with an additional condition to check that the difference between the two numbers is exactly 1
filtered_df_cn <- wid_data_cn[grepl(pattern, wid_data_cn$percentile) & 
                          as.numeric(sub("p(\\d{1,2})p(\\d{1,2})", "\\2", wid_data_cn$percentile)) -
                          as.numeric(sub("p(\\d{1,2})p(\\d{1,2})", "\\1", wid_data_cn$percentile)) == 1, ]

lorenz_curve <- ineq::Lc(filtered_df$value)

lorenz_curve_cn <- ineq::Lc(filtered_df_cn$value)



#### Calculate Gini:

# Step 1: Sort by income and create the rank (R) variable
filtered_df_cn <- filtered_df_cn[order(filtered_df_cn$value), ]
filtered_df_cn$rank <- seq_along(filtered_df_cn$value) / nrow(filtered_df_cn)

# Step 2: Calculate the mean of the income values (x̄)
mean_income <- mean(filtered_df_cn$value)

# Step 3: Calculate the covariance between income and rank
covar <- cov(filtered_df_cn$value, filtered_df_cn$rank)

# Step 4: Calculate the Gini coefficient using the formula
gini_coefficient_cn_formula <- 2/mean_income * covar
# 
# # Print the Gini coefficient
# print(gini_coefficient)

# Calculate the Gini coefficient using the ineq::Gini function
gini_coefficient_cn <- ineq::Gini(filtered_df_cn$value)

# Display the Gini coefficient
# print(gini_coefficient)
```

```{r}
#| label: fig-lorenz
#| fig-cap: "Lorenz curves for Sweden and Hungary"
#| fig-width: 5
#| fig-height: 3.9
#| fig-pos: H
## Graph Lorenz Curve:

plot(lorenz_curve,
     main = "Lorenz Curve",
     xlab = "Cumulative Share of Population",
     ylab = "Cumulative Share of Income",
     col = "blue", lwd = 2)

lines(lorenz_curve_cn, col = "red", lwd = 2)

legend("topleft",
       legend = c("Sweden", "Hungary", "Perfect equality"),
       col    = c("blue", "red", par("col")),
       lty    = c(1,     1,      1),
       lwd    = c(2,     2,      par("lwd")),
       bty = "n")
```

As shown in @fig-lorenz, the Lorenz curves for Sweden and Hungary indicate that Sweden's curve is closer to the perfect equality line than Hungary's, implying a more equal distribution in Sweden.

```{r}
#| label: tbl-gini
#| tbl-cap: "Gini coefficients for Sweden and Hungary"
#| echo: false
#| tbl-pos: H


library(dplyr)
library(gt)

gini_tbl <- tibble::tibble(
  Country = c("Sweden", "Hungary"),
  Gini_ineq   = c(gini_coefficient, gini_coefficient_cn),
  Gini_formula = c(gini_coefficient_formula, gini_coefficient_cn_formula)
)

gini_tbl |>
  gt() |>
  fmt_number(columns = c(Gini_ineq, Gini_formula), decimals = 3) |>
  cols_label(Country = "Country", Gini_ineq = "Gini - Ineq", Gini_formula = "Gini - formula") |> 
  cols_align(align = "center", columns = everything()) |>
  cols_width(
    Country ~ px(120),
    Gini_ineq    ~ px(140),
    Gini_formula ~ px(140)
  ) |> 
  tab_options(
    table.width = px(400),
    table.align = "center",
    data_row.padding = px(4),
    column_labels.padding = px(6)
  )
```

As reported in @tbl-gini, the Gini coefficient calculated by _Ineq_ package of Sweden is 0.364, slightly lower than Hungary's (0.383). This indicates that Sweden is more income-equal than Hungary.

However, the Gini coefficients computed via the formula and via the _Ineq_ package differ slightly. The formula-based calculation uses the sample covariance, which tends to be a bit larger, whereas the _Ineq_ package uses the population covariance, applying an $(n-1)/n$ adjustment to the formula-based result.

# Part 2

I pair Sweden with Hungary because Hungary is an interesting outlier among European countries---sometimes characterized as an "electoral autocracy,"[^1] suggesting distinctive institutional features. This contrast offers a useful case for assessing how intergenerational income elasticity differs across institutional contexts.

[^1]: https://www.europarl.europa.eu/news/en/press-room/20220909IPR40137

Based on $y_i=\alpha+\beta x+\delta z+\varepsilon$, I then decompose the variance of $y$ into components of $x$, $z$, and their covariance.

```{r}
#| results: hide
# Load necessary libraries
library(haven)
library(dplyr)

ess <- read_dta("~/Documents/IS/Lab 1/ESS9e03_2.dta")

# Filter for Sweden (SE) only
ess_se <- ess %>% filter(cntry == "SE")

# Convert eduyrs and edulvlb to numeric, coercing strings to NA
ess_se <- ess_se %>%
  mutate(
    eduyrs = as.numeric(eduyrs),
    edulvlb = as.numeric(edulvlb)
  )

# Filter out rows with NA values in eduyrs or edulvlb (including those that were strings)
ess_se_complete <- ess_se %>% filter(!is.na(eduyrs) & !is.na(edulvlb))

# Run the regression on the filtered dataset
model_edu <- lm(eduyrs ~ factor(edulvlb), data = ess_se_complete)

# Predict Eduyears for the filtered dataset
ess_se_complete <- ess_se_complete %>% mutate(Eduyears = predict(model_edu))

# Preserve the mapping: rename Eduyears to FatherEduyears
father_mapping <- ess_se_complete %>% 
  select(edulvlb, Eduyears) %>%
  rename(FatherEduyears = Eduyears, edulvlfb = edulvlb) %>%
  distinct()

# Merge the father education years into the main dataset based on edulvlfb
ess_se <- ess_se %>%
  left_join(father_mapping, by = c("edulvlfb" = "edulvlfb"))

# hinctnta <- log(hinctnta)

# Run the regression: hinctnta on eduyrs and FatherEduyears
model <- lm(hinctnta ~ eduyrs + FatherEduyears, data = ess_se)

# Results of regression:
summary(model)

# Extract coefficients
b_eduyrs <- coef(model)["eduyrs"]
b_FatherEduyears <- coef(model)["FatherEduyears"]

# Predict Residuals
#ess_se <- ess_se %>% mutate(Residuals = residuals(model))

# Calculate standard deviations
sd_eduyrs <- sd(ess_se$eduyrs, na.rm = TRUE)
sd_FatherEduyears <- sd(ess_se$FatherEduyears, na.rm = TRUE)
#sd_Residual <- sd(ess_se$Residuals, na.rm = TRUE)

# Calculate covariance between eduyrs and FatherEduyears
cov_eduyrs_FatherEduyears <- cov(ess_se$eduyrs, ess_se$FatherEduyears, use = "complete.obs")

# Calibration calculation
Calibration <- b_eduyrs^2 * sd_eduyrs^2 + 
  b_FatherEduyears^2 * sd_FatherEduyears^2 + 
  2 * b_eduyrs * b_FatherEduyears * cov_eduyrs_FatherEduyears

# Calculate the actual variance of hinctnta
sd_hinctnta <- sd(ess_se$hinctnta, na.rm = TRUE)
var_hinctnta <- sd_hinctnta^2

VarE_eduyrs <- b_eduyrs^2 * sd_eduyrs^2
VarE_FatherEduyears <- b_FatherEduyears^2 * sd_FatherEduyears^2
VarE_Cov <- 2 * b_eduyrs * b_FatherEduyears * cov_eduyrs_FatherEduyears

Calibration_r2 <- Calibration/var_hinctnta
```

```{r}
#| label: tbl-var-decomp-se
#| tbl-cap: "Variance decomposition results (Sweden)"
#| tbl-pos: H
#| echo: false

library(dplyr)
library(gt)

var_decomp_tbl <- tibble::tibble(
  Component = c(
    "Calibration",
    "Household income",
    "Years of education",
    "Father's years of education",
    "Covariance term"
  ),
  Value = c(
    Calibration,
    var_hinctnta,
    VarE_eduyrs,
    VarE_FatherEduyears,
    VarE_Cov
  )
)

var_decomp_share_tbl <- var_decomp_tbl |>
  mutate(Share = ifelse(Component == "Calibration", NA_real_, Value / var_hinctnta)) |>
  filter(Component != "Calibration")

var_decomp_share_tbl |>
  gt() |>
  fmt_markdown(columns = Component) |>
  fmt_number(columns = c(Value), decimals = 3) |>
  fmt_percent(columns = Share, decimals = 1) |>
  cols_label(
    Component = "Component",
    Value     = "Value",
    Share     = "Share of Total"
  ) |>
  cols_align(align = "center", columns = everything()) |>
  cols_width(
    Component ~ px(280),
    Value     ~ px(160),
    Share     ~ px(160)
  ) |>
  tab_options(
    table.width = px(620),
    table.align = "center",
    data_row.padding = px(4),
    column_labels.padding = px(6)
  ) |> 
  tab_source_note(
    md(sprintf("Calibration of $\\mathrm{R}^{2}$: %.3f", Calibration_r2))
  )
```
As shown in @tbl-var-decomp-se and @tbl-var-decomp-hu, the variance decompositions of income inequality in Sweden and Hungary differ markedly. In Sweden, years of education accounts for 8.4% of the variance in household income, father's education accounts for 0.9%, and the covariance term accounts for 1.9%. In Hungary, the respective shares are 4.1%, 4.7%, and 4.0%. These patterns indicate that individual's years of education is the dominant observed factor in Sweden, whereas in Hungary father's years of education and the covariance between individual's education and father's education account for a large share.
```{r}
#| results: hide
# Load necessary libraries
library(haven)
library(dplyr)

ess <- read_dta("~/Documents/IS/Lab 1/ESS9e03_2.dta")

# Filter for Sweden (SE) only
ess_se <- ess %>% filter(cntry == "HU")

# Convert eduyrs and edulvlb to numeric, coercing strings to NA
ess_se <- ess_se %>%
  mutate(
    eduyrs = as.numeric(eduyrs),
    edulvlb = as.numeric(edulvlb)
  )

# Filter out rows with NA values in eduyrs or edulvlb (including those that were strings)
ess_se_complete <- ess_se %>% filter(!is.na(eduyrs) & !is.na(edulvlb))

# Run the regression on the filtered dataset
model_edu <- lm(eduyrs ~ factor(edulvlb), data = ess_se_complete)

# Predict Eduyears for the filtered dataset
ess_se_complete <- ess_se_complete %>% mutate(Eduyears = predict(model_edu))

# Preserve the mapping: rename Eduyears to FatherEduyears
father_mapping <- ess_se_complete %>% 
  select(edulvlb, Eduyears) %>%
  rename(FatherEduyears = Eduyears, edulvlfb = edulvlb) %>%
  distinct()

# Merge the father education years into the main dataset based on edulvlfb
ess_se <- ess_se %>%
  left_join(father_mapping, by = c("edulvlfb" = "edulvlfb"))

# hinctnta <- log(hinctnta)

# Run the regression: hinctnta on eduyrs and FatherEduyears
model <- lm(hinctnta ~ eduyrs + FatherEduyears, data = ess_se)

# Results of regression:
summary(model)

# Extract coefficients
b_eduyrs <- coef(model)["eduyrs"]
b_FatherEduyears <- coef(model)["FatherEduyears"]

# Predict Residuals
#ess_se <- ess_se %>% mutate(Residuals = residuals(model))

# Calculate standard deviations
sd_eduyrs <- sd(ess_se$eduyrs, na.rm = TRUE)
sd_FatherEduyears <- sd(ess_se$FatherEduyears, na.rm = TRUE)
#sd_Residual <- sd(ess_se$Residuals, na.rm = TRUE)

# Calculate covariance between eduyrs and FatherEduyears
cov_eduyrs_FatherEduyears <- cov(ess_se$eduyrs, ess_se$FatherEduyears, use = "complete.obs")

# Calibration calculation
Calibration <- b_eduyrs^2 * sd_eduyrs^2 + 
  b_FatherEduyears^2 * sd_FatherEduyears^2 + 
  2 * b_eduyrs * b_FatherEduyears * cov_eduyrs_FatherEduyears

# Calculate the actual variance of hinctnta
sd_hinctnta <- sd(ess_se$hinctnta, na.rm = TRUE)
var_hinctnta <- sd_hinctnta^2

VarE_eduyrs <- b_eduyrs^2 * sd_eduyrs^2
VarE_FatherEduyears <- b_FatherEduyears^2 * sd_FatherEduyears^2
VarE_Cov <- 2 * b_eduyrs * b_FatherEduyears * cov_eduyrs_FatherEduyears

Calibration_r2 <- Calibration/var_hinctnta
```

```{r}
#| label: tbl-var-decomp-hu
#| tbl-cap: "Variance decomposition results (Hungary)"
#| tbl-pos: H
#| echo: false

library(dplyr)
library(gt)

var_decomp_tbl <- tibble::tibble(
  Component = c(
    "Calibration",
    "Household income",
    "Years of education",
    "Father's years of education",
    "Covariance term"
  ),
  Value = c(
    Calibration,
    var_hinctnta,
    VarE_eduyrs,
    VarE_FatherEduyears,
    VarE_Cov
  )
)

var_decomp_share_tbl <- var_decomp_tbl |>
  mutate(Share = ifelse(Component == "Calibration", NA_real_, Value / var_hinctnta)) |>
  filter(Component != "Calibration")

var_decomp_share_tbl |>
  gt() |>
  fmt_markdown(columns = Component) |>
  fmt_number(columns = c(Value), decimals = 3) |>
  fmt_percent(columns = Share, decimals = 1) |>
  cols_label(
    Component = "Component",
    Value     = "Value",
    Share     = "Share of Total"
  ) |>
  cols_align(align = "center", columns = everything()) |>
  cols_width(
    Component ~ px(280),
    Value     ~ px(160),
    Share     ~ px(160)
  ) |>
  tab_options(
    table.width = px(620),
    table.align = "center",
    data_row.padding = px(4),
    column_labels.padding = px(6)
  ) |> 
  tab_source_note(
    md(sprintf("Calibration of $\\mathrm{R}^{2}$: %.3f", Calibration_r2))
  )
```



The decomposition suggests that intergenerational income elasticity partially caused by father's education and its covariance with individual's education are more salient in Hungary. Institutional differences could underlie these contrasts. In other words, Sweden and Hungary differ in their intergenerational income inequality mechanisms.

Taken together, the Lorenz curves and Gini coefficients align with the variance decomposition: Sweden's lower overall inequality coincides with a larger role for individual's education, whereas the pattern in Hungary highlights father's education.