---
title: "How Nations Are Free: Exploring the Link Between Economic and Personal Freedoms"
author: "Yukun Jiao"
date: last-modified
date-format: "MMMM DD, YYYY"
format: pdf
header-includes:
  - \usepackage{setspace}
  - \onehalfspacing
editor: source
bibliography: references.bib
---

```{r}
#| include: false
library(tidyverse)
library(mice)
library(stringr)

# PCA, MCA
library(FactoMineR)
library(explor)

# Clustering
library(cluster)
library(NbClust)
library(factoextra)
library(ade4)

# t-SNE, SOM
library(Rtsne)
library(aweSOM)
library(Hmisc)
library(ggplot2)
```

```{r}
#| include: false
HFI <- read.csv("HFI/hfi2008_2016.csv")

supplementary_vars <- names(HFI)[sapply(str_split(names(HFI), "_"), function(x) length(x) == 1)] |> c("ISO_code")

rank_vars <- c("hf_rank", "hf_quartile", "ef_rank", "pf_rank")

level1_vars <- names(HFI)[sapply(str_split(names(HFI), "_"), function(x) length(x) == 2)] |> setdiff("ISO_code")

sub_vars <- names(HFI)[sapply(str_split(names(HFI), "_"), function(x) length(x) == 3)]

ssub_vars <- names(HFI)[sapply(str_split(names(HFI), "_"), function(x) length(x) == 4)]

sssub_vars <- names(HFI)[sapply(str_split(names(HFI), "_"), function(x) length(x) == 5)]
```

```{r}
#| include: false
########
score_and_rank <- names(HFI)[grep("rank|score|quartile", names(HFI))]

pf_vars <- names(HFI)[str_detect(names(HFI), "^pf_[^_]+$")] |> setdiff(score_and_rank)

ef_vars <- names(HFI)[str_detect(names(HFI), "^ef_[^_]+$")] |> setdiff(score_and_rank)

parent_vars <- sapply(str_split(sub_vars, "_"), function(parts) {
  paste(parts[1:2], collapse = "_")
})
names(parent_vars) <- sub_vars

mapping_df <- data.frame(
  sub_vars = names(parent_vars),
  parent_vars = parent_vars
)

selected_vars <- c(supplementary_vars, level1_vars, sub_vars)

for (i in 1:2) {
data <- 
  HFI |> 
  filter(!is.na(hf_score)) |> 
  select(all_of(selected_vars)) |> 
  mutate(across(all_of(pf_vars), ~ ifelse(is.na(.x), pf_score, .x))) |> 
  mutate(across(all_of(ef_vars), ~ ifelse(is.na(.x), ef_score, .x))) |> 
  mutate(across(all_of(mapping_df$sub_vars), ~ {
    cur_var <- cur_column()
    parent_var <- mapping_df$parent_vars[mapping_df$sub_vars == cur_var]
    if(length(parent_var) == 1 && parent_var %in% names(data)) {
    if_else(is.na(.x), data[[parent_var]], .x)
    } else {
      .x
    }
    }))
}

sum(is.na(data))
```

```{r}
#| message: false
#| warning: false
#| echo: false
selected_vars <- c(supplementary_vars, level1_vars)

# From 2008 to 2016
data_PCA <- data |> 
  select(all_of(selected_vars)) |> 
  select(-all_of(rank_vars))

freedom_scores <- c("year", "pf_score", "ef_score", "hf_score")

quanti_supp_score <- match(freedom_scores, colnames(data_PCA))

res <- PCA(data_PCA, quanti.sup = quanti_supp_score, quali.sup = c(2, 3, 4), graph = FALSE)

# explor(res)

res_explor <- explor::prepare_results(res)
PCA_plot_PC12 <- explor::PCA_var_plot(res_explor, xax = 1, yax = 2, var_sup = TRUE, var_sup_choice = c("pf_score",
    "ef_score", "hf_score"), var_lab_min_contrib = 0, col_var = "Type", labels_size = 10,
    scale_unit = TRUE, transitions = TRUE, labels_positions = NULL, xlim = c(-0.456,
        1.23), ylim = c(-0.704, 0.984))

PCA_plot_PC13 <- explor::PCA_var_plot(res_explor, xax = 1, yax = 3, var_sup = FALSE, var_sup_choice = ,
    var_lab_min_contrib = 0, col_var = "Type", labels_size = 10, scale_unit = TRUE,
    transitions = TRUE, labels_positions = NULL, xlim = c(-0.43, 1.08), ylim = c(-0.403,
        1.1))
```

```{r}
#| message: false
#| warning: false
#| echo: false
library(gt)
library(dplyr)
Variable_freedom <- c("Rule of Law", "Security and Safety", "Movement", "Religion", "Association, Assembly, and Civil Society", "Expression and Information", "Identity and Relationships", "Size of Government", "Legal System and Property Rights", "Access to Sound Money", "Freedom to Trade Internationally", "Regulation of Credit, Labor, and Business")

df_PCA <- as.data.frame(res$var$coord[, 1:4])
df_PCA$Variable <- Variable_freedom
df_PCA <- df_PCA |>
  select(Variable, everything())
eigenvalues <- res$eig[1:4, 1]
percentages <- res$eig[1:4, 2]

eigen_row <- data.frame(
  Variable = "Eigenvalue",
  Dim.1 = eigenvalues[1],
  Dim.2 = eigenvalues[2],
  Dim.3 = eigenvalues[3],
  Dim.4 = eigenvalues[4]
)
percentage_row <- data.frame(
  Variable = "Percentage (%)",
  Dim.1 = percentages[1],
  Dim.2 = percentages[2],
  Dim.3 = percentages[3],
  Dim.4 = percentages[4]
)

df_PCA <- bind_rows(df_PCA, eigen_row, percentage_row)

gt_table <- gt(df_PCA) |> 
  fmt_number(
    columns = 2:5,
    decimals = 3
  ) |> 
  fmt_number(
    rows = 14,
    columns = 2:5,
    decimals = 1
  ) |> 
  cols_label(
    Variable = "Variable",
    Dim.1 = "PC1",
    Dim.2 = "PC2",
    Dim.3 = "PC3",
    Dim.4 = "PC4"
  )
```

```{r}
#| message: false
#| warning: false
#| echo: false

data_PCA_2016 <- data |> 
  filter(year == 2016) |> 
  select(all_of(selected_vars)) |> 
  select(-all_of(rank_vars))
res_2016 <- PCA(data_PCA_2016, quanti.sup = quanti_supp_score, quali.sup = c(2, 3, 4), graph = FALSE)

pc_scores <- as.data.frame(res_2016$ind$coord[, 1:4])

pc_scores$countries <- data_PCA_2016$countries
pc_scores <- pc_scores |> 
  select(countries, everything())

tbl_score <- function(dim) {
  top_10 <- pc_scores |> 
    arrange(desc({{dim}})) |> 
    slice_head(n = 10) |> 
    mutate(Group = "Top 10")
  
  bottom_10 <- pc_scores |> 
    arrange({{dim}}) |> 
    slice_head(n = 10) |> 
    mutate(Group = "Bottom 10")
  
  top_bottom_10 <- bind_rows(top_10, bottom_10)
  
  gt_tbl_scores <- top_bottom_10 |> 
    select(Group, countries, Dim.1, Dim.2, Dim.3, Dim.4) |> 
    gt(groupname_col = "Group")|> 
    fmt_number(
      columns = 2:5,
      decimals = 3
    ) |> 
    cols_label(
      Dim.1 = "PC1",
      Dim.2 = "PC2",
      Dim.3 = "PC3",
      Dim.4 = "PC4"
    )
return(gt_tbl_scores)
}

```

```{r}
#| message: false
#| warning: false
#| echo: false

selected_vars <- c(supplementary_vars, level1_vars)

# From 2008 to 2016
# data_SOM <- data |> 
#   select(all_of(selected_vars)) |> 
#   select(-all_of(score_and_rank)) |> 
#   select(-all_of(supplementary_vars)) |> 
#   scale()
# 
# 
# countries_year <- paste0(data$countries, "_", data$year)
# rownames(data_SOM) <- countries_year

# Only for 2016
selected_vars <- c(supplementary_vars, level1_vars)
countries_info <- data[data$year == 2016, "countries"]
region_info <- data[data$year == 2016, "region"]

data_SOM <- data |> 
  filter(year == 2016) |> 
  select(all_of(selected_vars)) |> 
  select(-all_of(score_and_rank)) |> 
  select(-all_of(supplementary_vars)) |> 
  scale()

# res_SOM <- PCA(data_SOM)
# 
# fviz_nbclust(data_SOM, hcut, hc_method = "ward.D2", method = c("silhouette"))

rownames(data_SOM) <- countries_info
write.csv(data_SOM, "data_SOM.csv")
# aweSOM()
```


# Abstract

本研究使用PCA和SOM to investigate the HFI data from 2008 to 2016 to 提升对数据集本身的理解. 并提出了两个相关的探索性的问题。

The result of PCA indicates that the first principal component corresponds to a measure of the human freedom; the second principal component corresponds to a measure of authoritarian capitalism. the third principal component corresponds to a measure of social democracy. It also suggests that rule of law is most important factor contributing to both personal freedom and economic freedom, which is consistent with previous literatures.

The result of SOM suggests the same interpretation as PCA did. But It provides extra clustering information for us to understanding the underlying patterns among the data. It suggests that 对于已有国家数据而言，大致存在四种国家分类，分别是社会民主主义国家，法治较差的资本主义国家，威权资本主义国家和战乱状态下的国家（在数据上表现为人类自由的全面缺失）。

These exploratory results contribute to explore the underlying factors that enable certain countries to advance economic freedom despite limited personal freedom, 这为我们理解不同的发展模式提供了帮助.

# Introduction

Friedman points out that there is an intimate connection between economics and politics, and only certain combinations of political and economic arrangements are possible. In particular, a society that is socialist cannot also be democratic in the sense of guaranteeing individual freedom. Economic arrangements play a dual role in promoting a free society: economic freedom is both a component of freedom broadly understood as an end in itself and an indispensable means toward achieving political freedom.

Building upon the understanding of economic and political freedom, it is important to examine the role of the free market and government in sustaining these freedoms. Friedman argued that the role of competitive capitalism the organization of the bulk of economic activity through private enterprise operating in a free market as a system of economic freedom and a necessary condition for political freedom. The competitive capitalism plays a crucial role in limiting the concentration of power by separating economic power from political power. Even in societies where political freedom is lacking, such as Fascist regimes or tsarist Russia, the presence of competitive capitalism and private property provided a check on centralized state power, allowing citizens more freedom than those in totalitarian states where economic and political totalitarianism are combined. Thus, a free market can help constrain government authority and preserve some degree of individual freedom even when political institutions are not fully free [@friedman1982]. In other words, economic freedom can greatly reduce the range of issues that must be decided through political means, thereby minimizing direct government involvement and maximizing personal freedom.

Structural characteristics such as low national income, proximity to the equator, high ethnolinguistic diversity, French or socialist legal origins, and a high proportion of Catholics or Muslims have been found to correlate with inferior government performance [@lopez-de-silanes1999]. Although these factors are associated with lower government performance rather than freedom directly, ineffective governance often limits the protection of rights, the enforcement of contracts, and the overall rule of law—all of which are essential for both personal and economic freedom. In fact, a well-functioning government can support freedom. The generally positive or neutral relationship between government activity and economic freedom suggests that there is no inherent contradiction between the two. Moreover, free markets depend on institutional quality and credible oversight—functions that governments are well-positioned to provide [@ott2018].

Institutional frameworks, especially the rule of law, play a fundamental role in supporting both economic growth and freedom. Over the past decades, scholars have recognized the crucial role of the rule of law in promoting economic growth. By constraining arbitrary governmental actions, the rule of law creates micro-level incentives and provides an institutional framework conducive to investment, entrepreneurship, and long-term capital development. Empirical evidence robustly shows that stability in legal rules, secure property rights, and restraint on arbitrary government actions significantly encourage economic growth. The varying economic outcomes among countries are largely explained by differences in institutional quality, with the rule of law being a key distinguishing factor. In contrast, democratic elections appear far less important for economic growth, as democracy does not necessarily generate the institutions needed for sustained growth and may instead foster redistributive and special interest politics that hinder it. This explains why countries such as Chile, Singapore, and Hong Kong have experienced rapid economic growth despite lacking democratic politics; strong rule of law values in these countries provided an environment conducive to economic prosperity [@zywicki2003].

Building on the theoretical foundations of economic and personal freedom, and recognizing the critical role of institutional quality in supporting these freedoms, it becomes essential to develop rigorous empirical methods to measure and analyze the complex dimensions of human freedom. Among various metrics, the widely cited Human Freedom Index (HFI) provides a comprehensive framework for assessing freedom, yet its constituent dimensions remain insufficiently examined. Previous studies have applied principal component analysis, k-means clustering, and factor analysis to investigate the structure of the Human Freedom Index (HFI), yielding valuable insights [@murphy2025]. Building on previous work, this study offers a complementary perspective by applying PCA with a focus on interpretability and dimensional clarity.

Based on previous review, this study focuses on the following exploratory questions:

RQ1: What factors allow certain countries to advance economic freedom despite limited personal freedom?

RQ2: In the long run, what factors could enable these countries to transition into states with high levels of personal freedom (rule of law and specific personal freedom)?

# Data

The Human Freedom Index (HFI) presents a broad measure of human freedom, understood as the absence of coercive constraint [@ian2018]. The index is constructed using 79 distinct indicators that capture both personal and economic dimensions of freedom. These indicators span a wide range of 12 areas, including the rule of law, security and safety, freedom of movement, religion, association, assembly, and civil society, as well as freedom of expression and information, identity and relationships, the size of government, the legal system and property rights, access to sound money, freedom to trade internationally, and the regulation of credit, labor, and business.

The HFI is one of the most comprehensive freedom indices constructed for a globally representative set of countries and jurisdictions. The 2018 edition of the index covers 162 jurisdictions and provides annual data from 2008 to 2016. This study uses the 2018 edition as its data source, with freedom measured on a scale of 0 to 10, where 10 represents more freedom. In the data processing of this study, countries with missing Human Freedom Score were excluded; for the remaining observations with missing indicators, the missing values were imputed based on the hierarchical aggregation structure of the index, using the value of the indicator at the higher level.

# Method
## Principal Component Analysis (PCA)

The goal of PCA is to find a low-dimensional representation of a data set that contains as much as possible of the variation. For n observations and p features, each of the dimensions found by PCA is a linear combination of the p features. These dimensions are the principal components we seek. The first principal component is the normalized linear combination of the features that has the largest variance. The second principal component is the linear combination of $X_1$, \ldots, $X_p$ that has the maximal variance among all linear combinations that are uncorrelated with the first principal component. And so on. That is, PCA finds principal components that are orthogonal to one another, with each having the maximum possible variance.

Once the principal components are computed, we can generate a low-dimensional view of the data. Typically, such a view shows both the principal component scores and the loading vectors. The loading vectors indicate directions in the feature space along which the data vary the most, and the scores are projections of the observations onto these directions. Moreover, the contribution of a variable helps interpret the meaning of each principal component, and the $cos^2$ of a variable can be used to measure the quality of its representation.

A limitation of PCA is that it does not necessarily reflect the practical or theoretical importance of individual variables. Specifically, a variable with a high weight on a principal component merely indicates a strong contribution to explaining variance, but this does not necessarily mean that the variable is more important in practical or theoretical terms.

## Self-Organizing Map (SOM)

SOM is an unsupervised machine learning algorithm based on competitive learning [@kohonen1990]. The SOM represents a distribution of input data items using a finite set of models, in which these models are automatically associated with the nodes of a regular grid in an orderly fashion such that more similar models become automatically associated with nodes that are adjacent in the grid, whereas less similar models are situated farther away from each other in the grid. This organization makes it possible to obtain an insight into the topographic relationships of data, especially of high-dimensional data items. The central idea of SOM is that every input data item shall select the model that matches best with the input item, and this model, as well as a subset of its spatial neighbors in the grid, shall be modified for better matching. The quality of SOM results can be evaluated using quantization error, which measures the average distance between data points and their corresponding map nodes (with smaller values indicating better representation), and topographic error, which assesses how well the SOM preserves the topological structure of the input space by identifying local discontinuities.

To create a SOM, the data first need to be scaled, and then the number of cells must be determined. However, it is not possible to estimate the exact size of the cells in advance. It must be determined by the trial-and-error method, after seeing the quality of the first guess. Typical SOM arrays range from a few dozen to a few hundred nodes, and the relation of the horizontal and vertical dimensions of the array ought to comply at least roughly with the relation of the two largest principal components of the input data, respectively. The number of cells could be determined by a heuristic formula, $M \approx 5\sqrt{N}$, where $M$ is the number of neurons, which is an integer close to the result of the right hand side of the equation, and $N$ is the number of observations [@tian2014]. Given $N = 162$, this yields approximately 64 neurons. Moreover, to preserve the topological structure of the input data, the aspect ratio of the SOM map is chosen to reflect the approximate ratio between the first and second principal components ($\text{PC}_1 / \text{PC}_2 \approx 3$). Based on the above considerations and the results of multiple trials, a $4 \times 12$ map (4 rows and 12 columns) was finally selected.

Moreover, SOM is often used as a kind of histogram on which one displays the number of input data items that is mapped into each of the nodes. In this case, a very coarse rule-of-thumb may be that about 50 items per cell on the average should be sufficient, otherwise the resolution is limited by the sparsity of data [@kohonen2013].

# Results

In the HFI data from 2008 to 2016, the first principal component explains 47.7% of the variance in the data, the second principal component explains 14.7% of the variance in the data, and the third principal component explains 9.8% of the variance. Together, the first three principal components explain almost 72.2% of the variance in the data, and the fourth principal components explain only 5.8% of the variance. And The fourth principal components does not show interesting pattern. Moreover, there is an elbow after the third component as shown in [@fig-scree].

Specifically, as shown in [@tbl-PCA], the first principal component loads positively on all variables except the size of government, and it corresponds to a measure of human freedom.

The second principal component places approximately equal weight on all personal freedoms except rule of law and security and safety---which freedoms are civil freedoms---and also places a small positive weight on the size of government. It also places negative weight on all economic freedoms except size of government. This means it seems to correspond to a measure of civil freedom or economic freedom on the other side. Because principal components are orthogonal, this implies that in countries that already guarantee relative freedom, civil freedom and economic freedom are opposed, which sounds strange because economic freedom should be a means to realize personal freedom, not an opposing relationship. Therefore, the correct explanation should be that the second component corresponds to a form of authoritarian capitalism, in which the government promotes economic freedom while suppressing personal freedom. This may cause the data to show such a pattern.

The third principal component places most weight on size of government and roughly equal positive weights on several economic freedoms and roughly equal negative weights on several personal freedoms. This loading vector should roughly correspond to a measure of Western European social democracy, except several countries with very large government or very small government.

The fourth principle component loads mainly on identity and relationship. But its Eigenvalue and not so interesting patterns suggests that this loading vector may not be particularly meaningful.

Although the second principal component does not account for the largest amount of variance, its pattern is the most interesting. Therefore, the top and bottom 10 countries by PC2 scores for 2016 are listed separately, as shown in [@tbl-PC2]. The top 10 countries suggest a pattern characterized by a lack of economic freedom. the bottom 10 countries reflect a pattern of relatively high levels of economic freedom, but with notable restrictions on specific personal freedom compared to other democratic countries with similar economic freedom.

As shown in [@fig-screeSOM], there is a clear elbow at the point where the number of superclasses reaches 4. While the silhouette of cell superclasses in [@fig-silhouette] was also considered, its contribution is less pronounced. In fact, when the number of superclasses is less than or equal to 3, the average silhouette width remains high and stable, indicating generally good and consistent clustering performance. However, such a small number of superclasses offers limited explanatory power. After multiple trials and evaluation, the number of superclasses was set to 4 to allow the SOM to achieve the most interpretable clustering outcome. According to [@tbl-Quality_SOM], this map explained 82.79% variance and remained the data topographic structure very well.

Then, the observation cloud of the SOM is shown in [@fig-SOM]. The clustering pattern reveals the following groups: The lower-right corner is dominated by Western European social democracies, such as Sweden and Denmark; The upper-right corner clusters Hong Kong and Singapore, representing distinctive economic models with strong rule of law; The lower-left corner gathers authoritarian capitalist regimes with relatively strong rule of law, including China and Russia. The upper-left corner consists of countries with severely restricted personal and economic freedoms, likely due to ongoing conflicts or wars and extremely weak rule of law. The central yellow cells gather capitalist nations with poorer performance, characterized by weak rule of law and weak legal system and property right.

Based on [@fig-histogram] and [@fig-radar], a more detailed explanation can be provided. This map corresponds roughly to the interpretation of the three principal components from the previous PCA. With the cell in the lower-left corner established as the origin of a Cartesian coordinate system, where Axis 1 runs from left to right and Axis 2 from bottom to top, Axis 1 represents human freedom. From left to right, a contrast is shown between unfree countries and free countries. Axis 2 corresponds to a measure of authoritarian capitalism and Western European social democracy. From bottom to top, an opposition is reflected between civil and economic freedoms as well as between social democracy and liberalism (Axis 2 does not apply to countries in the upper-left corner).


Based on previous exploratory results, the following hypothesis is proposed:

H1: The rule of law is the most influential factor affecting freedom.

The rule of law is the foundation of economic freedom. In the process of fostering political freedom, economic freedom requires the rule of law. A sound legal system ensures that the government maintains economic activities and safeguards economic freedom.

The rule of law is also the cornerstone of individual freedom. It limits arbitrary actions by the government, thereby protecting individual rights from being infringed upon by public authority. At the same time, legal norms prevent violations of individual rights by private actors.

H2: Civil liberties (specific personal freedoms) are not a necessary condition for economic growth.

In promoting economic growth, the importance of democratic elections is far less than that of the rule of law itself. In fact, democracy may lean toward redistributive politics and the competition of special interest groups, which often suppress economic growth.

H3: Economic freedom, accompanied by market economic activities, disperses governmental power and weakens centralized authority, leading to a balance of power between economic and political forces.

# Conclusion

In the future,this study will further explore the distinguishing characteristics of countries where economic liberalization leads to political liberalization, as opposed to those where such a transition does not occur.

# References

::: {#refs}
:::


# Appendix

```{r}
#| message: false
#| warning: false
#| echo: false
#| label: fig-PC12
#| fig-cap: Variables Plot along PC1 and PC2 of PCA, 2008--2016

PCA_plot_PC12
```

```{r}
#| message: false
#| warning: false
#| echo: false
#| label: fig-PC13
#| fig-cap: Variables Plot along PC1 and PC3 of PCA, 2008--2016

PCA_plot_PC13
```

```{r}
#| message: false
#| warning: false
#| echo: false
#| label: fig-scree
#| fig-cap: Eigenvalues Histogram (Scree Plot) of PCA, 2008--2016

fviz_eig(res, addlabels = TRUE) +
  labs(x = "Principal Components",
       y = "Percentage of Variance Explained")
```


```{r}
#| message: false
#| warning: false
#| echo: false
#| label: tbl-PCA
#| tbl-cap: The First Four Principal Components of the Human Freedom Index, 2008--2016

gt_table
```

```{r}
#| message: false
#| warning: false
#| echo: false
#| label: tbl-PC2
#| tbl-cap: Top and Bottom 10 Countries by PC2 Score, 2016

  top_10 <- pc_scores |> 
    arrange(desc(Dim.2)) |> 
    slice_head(n = 10) |> 
    mutate(Group = "Top 10")
  
  bottom_10 <- pc_scores |> 
    arrange(Dim.2) |> 
    slice_head(n = 10) |> 
    mutate(Group = "Bottom 10")
  
  top_bottom_10 <- bind_rows(top_10, bottom_10)
  
  gt_tbl_scores <- top_bottom_10 |> 
    select(Group, countries, Dim.2) |> 
    gt(groupname_col = "Group")|> 
    fmt_number(
      columns = 3,
      decimals = 3
    ) |> 
    cols_label(
      Dim.2 = "PC2"
    )

gt_tbl_scores
```


```{r}
#| message: false
#| warning: false
#| echo: false

library(aweSOM) # (version 1.3) 

## Import Data
# setwd("/path/to/datafile/directory") ## Uncomment this line and set the path to the datafile's directory
import.data <- data.table::fread("data_SOM.csv", stringsAsFactors = TRUE, data.table = FALSE, check.names = TRUE)

## Build training data (numeric)
train.data <- import.data[, c("pf_rol", "pf_ss", "pf_movement", "pf_religion", "pf_association", "pf_expression", "pf_identity", "ef_government", "ef_legal", "ef_money", "ef_trade", "ef_regulation")]
### Scale training variables (unit variance)
train.data <- scale(train.data)
### Prepare plotting data
plot.data <- import.data

## Train SOM
### RNG Seed (for reproducibility)
set.seed(23968)
### Initialization
init <- somInit(train.data, ncols = 12, nrows = 4)
### Training
the.som <- kohonen::som(train.data, grid = kohonen::somgrid(12, 4, "hexagonal"), maxNA.fraction = 0.25, rlen = 100, alpha = c(0.05, 0.01), radius = c(5.29,-5.29), init = init, dist.fcts = "sumofsquares")

## Quality measures
# somQuality(the.som, train.data)


## Group cells into superclasses (hierarchical clustering)
superclust <- hclust(dist(the.som$codes[[1]]), "complete")
superclasses <- cutree(superclust, 4)
## Apply clusterings to observations
obs.class <- the.som$unit.classif
obs.superclass <- superclasses[obs.class]

```

```{r}
#| message: false
#| warning: false
#| echo: false
#| label: fig-screeSOM
#| fig-cap: SOM Superclass Scree plot, 2016
aweSOMscreeplot(the.som, method = "hierarchical", hmethod = "complete", nclass = 4)
```


```{r}
#| message: false
#| warning: false
#| echo: false
#| label: tbl-Quality_SOM
#| tbl-cap: SOM Quality Measures, 2016

Quality_som <- somQuality(the.som, train.data)

Quality_SOM <- list(
  err.quant = Quality_som$err.quant,
  err.varratio = Quality_som$err.varratio,
  err.topo = Quality_som$err.topo,
  err.kaski = Quality_som$err.kaski
)


quality_df <- data.frame(
  Measure = c(
    "Quantization Error",
    "Explained Variance (%)",
    "Topographic Error",
    "Kaski-Lagus Error"
  ),
  Value = c(
    Quality_som$err.quant,
    Quality_som$err.varratio,
    Quality_som$err.topo,
    Quality_som$err.kaski
  )
)

quality_table <- gt(quality_df) |> 
  fmt_number(
    columns = vars(Value),
    decimals = 2
  )

quality_table

```



```{r}
#| message: false
#| warning: false
#| echo: false
#| label: fig-silhouette
#| fig-cap: Silhouette of Cell Superclasses, 2016


# This one doesn't work as R script in R studio.
# aweSOMplot(som = the.som, type = "Hitmap", 
#            superclass = superclasses, obsNames = plot.data[, "V1"], 
#            size = 1000)

aweSOMsilhouette(the.som, superclasses)


# This one also doesn't work as R script in R studio.
# aweSOMplot(som = the.som, type = "UMatrix", 
#            superclass = superclasses, obsNames = plot.data[, "V1"], 
#            legendFontsize = 34,
#            size = 1000)

```

```{r}
#| message: false
#| warning: false
#| echo: false
#| label: fig-SOM
#| fig-cap: SOM Observation Cloud, 2016
## Interactive plot
aweSOMplot(som = the.som, type = "Cloud", data = plot.data, 
           variables = c("None", "pf_rol", "pf_ss", "pf_movement", "pf_religion", "pf_association", "pf_expression", "pf_identity", "ef_government", "ef_legal", "ef_money", "ef_trade", "ef_regulation"),
           superclass = superclasses, obsNames = plot.data[, "V1"], 
           legendFontsize = 32,
           size = 1000)
```

```{r}
#| message: false
#| warning: false
#| echo: false
#| label: fig-histogram
#| fig-cap: SOM Barplot of Numeric Variables, 2016
## Interactive plot
aweSOMplot(som = the.som, type = "Barplot", data = plot.data, 
           variables = c("pf_rol", "pf_ss", "pf_movement", "pf_religion", "pf_association", "pf_expression", "pf_identity", "ef_government", "ef_legal", "ef_money", "ef_trade", "ef_regulation"),
           superclass = superclasses, obsNames = plot.data[, "V1"], 
           legendFontsize = 34,
           size = 1000)

```
```{r}
#| message: false
#| warning: false
#| echo: false
#| label: fig-radar
#| fig-cap: SOM Radar Chart of Numeric Variables, 2016
## Interactive plot
aweSOMplot(som = the.som, type = "Radar", data = plot.data, 
           variables = c("pf_rol", "pf_ss", "pf_movement", "pf_religion", "pf_association", "pf_expression", "pf_identity", "ef_government", "ef_legal", "ef_money", "ef_trade", "ef_regulation"),
           superclass = superclasses, obsNames = plot.data[, "V1"], 
           legendFontsize = 34,
           size = 1000)

```

