---
title: "Inequality and Segregation Lab 3"
author: "Yukun Jiao"
date: "2025-12-01"
toc: false
format:
  html: 
    embed-resources: true
knitr: 
  opts_chunk:
    fig.align: center
editor: 
  markdown: 
    wrap: 72
execute:
  echo: true
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(ggforce)
library(cowplot)
library(RANN)
library(sf)
library(gt)
library(knitr)
library(kableExtra)

make_housing<-function(dim=100,n=10000,foci=c(0.2,0.8),noise=1,pow=2) {
  xy <-bind_rows(
    expand_grid( x=seq(0,dim)/dim, y=seq(0.5,dim-0.5)/dim ) ,
    expand_grid( x=seq(0.5,dim-0.5)/dim , y=seq(0,dim)/dim) )
  if(n>=nrow(xy)) {
    return(xy)
  }
  else {
    foci<-as_tibble(expand.grid(x_f=foci,y_f=foci))
    places<-expand_grid(xy,foci) |> 
      mutate(d=sqrt((x-x_f)^2+(y-y_f)^2)) |> 
      group_by(x,y) |> 
      summarize(u=sum(1/(d+0.001)^pow)) |> 
      ungroup() |> 
      mutate(u=u/sd(u)) |> 
      mutate(v = u + rnorm(n(),0,sd=noise)) |> 
      filter(v>=sort(v,decreasing=T)[n]) |> 
      select(x,y)
    return(places)
  }
}
```


# Modifying the Segregation Scenarios

## reducing to a single focus.
```{r}
#| warning: false

foci<-as_tibble(expand.grid(x_f=c(0.5),y_f=c(0.5)))


places<-make_housing(pow=1,noise=0.5,foci=c(0.5),n=10000)
ggplot() +
  geom_point(data=places,aes(x=x,y=y),size=0.1) +
  geom_point(data=foci,aes(x=x_f,y=y_f),color="green3",size=5,shape=4,stroke=3) +
  theme_void() + coord_fixed()

places<-places |> 
  expand_grid(foci) |> 
  mutate(d=sqrt((x-x_f)^2+(y-y_f)^2)) |> 
  mutate(u=1/d) |> 
  group_by(x,y) |> 
  summarize(u=sum(u)) |> 
  ungroup() |> # do this so calculations aren't performed within combos of x & y
  mutate(C=u+rnorm(n=length(u),mean=0,sd=sd(u))) |> 
  mutate(C=factor(ifelse(C>median(C),"red","blue"))) |> 
  ungroup() |> 
  mutate(A=u+rnorm(n=length(u),mean=0,sd=0*sd(u))) |>  # no noise!
  mutate(A=factor(ifelse(A>median(A),"red","blue"))) |> 
  mutate(B=u+rnorm(n=length(u),mean=0,sd=0.2*sd(u))) |> # a bit of noise
  mutate(B=factor(ifelse(B>median(B),"red","blue"))) |> 
  mutate(D=u+rnorm(n=length(u),mean=0,sd=3*sd(u))) |> # a lot of noise
  mutate(D=factor(ifelse(D>median(D),"red","blue")))


places <- places |> pivot_longer(cols=c("A","B","C","D"),
               names_to="city",
               values_to="group")

ggplot(data=slice(places,sample(1:nrow(places))), # random plot order
       aes(x=x,y=y,color=group)) + 
  geom_point(size=0.1) + 
  scale_color_identity() +
  coord_fixed() +
  theme_void() + 
  theme(legend.position = "none", aspect.ratio = 1) + 
  facet_grid(cols=vars(city)) -> gg_points

# gg_points

grid_dim<-10
blocks<-places |> 
  mutate(x=ceiling(x*grid_dim)/grid_dim,y=ceiling(y*grid_dim)/grid_dim) |> 
  group_by(city,x,y) |> 
  summarize(red=sum(ifelse(group=="red",1,0)),
            blue=sum(ifelse(group=="blue",1,0))) |> 
  mutate(`% red`=100*red/(red+blue))

ggplot(blocks,aes(x=x,y=y,fill=`% red`)) + 
  geom_raster() + 
  facet_grid(cols=vars(city)) + 
  coord_fixed() +
  scale_fill_gradient2(midpoint=50,high="red",mid="white",low="blue") + 
  theme_void() ->
  gg_blocks

plot_grid(gg_points,gg_blocks,nrow=2,align="hv",axis="r")
```

## Increasing the number of foci.
```{r}
#| warning: false

foci<-as_tibble(expand.grid(x_f=c(0.1, 0.5, 0.75),y_f=c(0.1, 0.5, 0.75)))

places<-make_housing(pow=1,noise=0.5,foci=c(0.1, 0.5, 0.75),n=10000)
ggplot() +
  geom_point(data=places,aes(x=x,y=y),size=0.1) +
  geom_point(data=foci,aes(x=x_f,y=y_f),color="green3",size=5,shape=4,stroke=3) +
  theme_void() + coord_fixed()

places<-places |> 
  expand_grid(foci) |> 
  mutate(d=sqrt((x-x_f)^2+(y-y_f)^2)) |> 
  mutate(u=1/d) |> 
  group_by(x,y) |> 
  summarize(u=sum(u)) |> 
  ungroup() |> # do this so calculations aren't performed within combos of x & y
  mutate(C=u+rnorm(n=length(u),mean=0,sd=sd(u))) |> 
  mutate(C=factor(ifelse(C>median(C),"red","blue"))) |> 
  ungroup() |> 
  mutate(A=u+rnorm(n=length(u),mean=0,sd=0*sd(u))) |>  # no noise!
  mutate(A=factor(ifelse(A>median(A),"red","blue"))) |> 
  mutate(B=u+rnorm(n=length(u),mean=0,sd=0.2*sd(u))) |> # a bit of noise
  mutate(B=factor(ifelse(B>median(B),"red","blue"))) |> 
  mutate(D=u+rnorm(n=length(u),mean=0,sd=3*sd(u))) |> # a lot of noise
  mutate(D=factor(ifelse(D>median(D),"red","blue")))

places <- places |> pivot_longer(cols=c("A","B","C","D"),
               names_to="city",
               values_to="group")

ggplot(data=slice(places,sample(1:nrow(places))), # random plot order
       aes(x=x,y=y,color=group)) + 
  geom_point(size=0.1) + 
  scale_color_identity() +
  coord_fixed() +
  theme_void() + 
  theme(legend.position = "none", aspect.ratio = 1) + 
  facet_grid(cols=vars(city)) -> gg_points

# gg_points


grid_dim<-10
blocks<-places |> 
  mutate(x=ceiling(x*grid_dim)/grid_dim,y=ceiling(y*grid_dim)/grid_dim) |> 
  group_by(city,x,y) |> 
  summarize(red=sum(ifelse(group=="red",1,0)),
            blue=sum(ifelse(group=="blue",1,0))) |> 
  mutate(`% red`=100*red/(red+blue))

ggplot(blocks,aes(x=x,y=y,fill=`% red`)) + 
  geom_raster() + 
  facet_grid(cols=vars(city)) + 
  coord_fixed() +
  scale_fill_gradient2(midpoint=50,high="red",mid="white",low="blue") + 
  theme_void() ->
  gg_blocks

plot_grid(gg_points,gg_blocks,nrow=2,align="hv",axis="r")
```
## Changing the spatial distribution of the foci.
```{r}
#| warning: false

foci<-as_tibble(expand.grid(x_f=c(0.1),y_f=c(0.1)))

places<-make_housing(pow=1,noise=0.5,foci=c(0.1),n=10000)
ggplot() +
  geom_point(data=places,aes(x=x,y=y),size=0.1) +
  geom_point(data=foci,aes(x=x_f,y=y_f),color="green3",size=5,shape=4,stroke=3) +
  theme_void() + coord_fixed()

places<-places |> 
  expand_grid(foci) |> 
  mutate(d=sqrt((x-x_f)^2+(y-y_f)^2)) |> 
  mutate(u=1/d) |> 
  group_by(x,y) |> 
  summarize(u=sum(u)) |> 
  ungroup() |> # do this so calculations aren't performed within combos of x & y
  mutate(C=u+rnorm(n=length(u),mean=0,sd=sd(u))) |> 
  mutate(C=factor(ifelse(C>median(C),"red","blue"))) |> 
  ungroup() |> 
  mutate(A=u+rnorm(n=length(u),mean=0,sd=0*sd(u))) |>  # no noise!
  mutate(A=factor(ifelse(A>median(A),"red","blue"))) |> 
  mutate(B=u+rnorm(n=length(u),mean=0,sd=0.2*sd(u))) |> # a bit of noise
  mutate(B=factor(ifelse(B>median(B),"red","blue"))) |> 
  mutate(D=u+rnorm(n=length(u),mean=0,sd=3*sd(u))) |> # a lot of noise
  mutate(D=factor(ifelse(D>median(D),"red","blue")))

# ggplot(places,aes(x=x,y=y,color=C)) +
#   geom_point() + 
#   scale_color_identity() +
#   theme_void() + 
#   coord_fixed() + 
#   theme(legend.position = "none")

places <- places |> pivot_longer(cols=c("A","B","C","D"),
               names_to="city",
               values_to="group")

ggplot(data=slice(places,sample(1:nrow(places))), # random plot order
       aes(x=x,y=y,color=group)) + 
  geom_point(size=0.1) + 
  scale_color_identity() +
  coord_fixed() +
  theme_void() + 
  theme(legend.position = "none", aspect.ratio = 1) + 
  facet_grid(cols=vars(city)) -> gg_points
gg_points
```


<!-- -   Switching to a linear assignment function. -->
<!-- -   Applying a logarithmic or some other non-linear assignment function. -->
<!-- -   Increasing the share of red dots. -->
<!-- -   Decreasing the share of red dots. -->
<!-- -   CHALLENGE: Making the *locations* (*not* a particular group) denser around the foci. -->




# Evenness and exposure measures of segregation

## Confirm that the index of dissimilarity obeys composition invariance principle by doubling the size of the "red" population in each neighborhood and re-calculating the indices.
```{r}
calcD <- function(x,y) {
  result<-0.5*sum(abs(x/sum(x)-y/sum(y)))
  return(result)
}

grid_dim<-10
blocks<-places |> 
  mutate(x=ceiling(x*grid_dim)/grid_dim,y=ceiling(y*grid_dim)/grid_dim) |> 
  group_by(city,x,y) |> 
  summarize(red=sum(ifelse(group=="red",1,0)),
            blue=sum(ifelse(group=="blue",1,0))) |> 
  mutate(`% red`=100*red/(red+blue))


D0 <- blocks |> 
  group_by(city) |> 
  summarize(D = calcD(red, blue), .groups="drop")

blocks_red2 <- blocks |> mutate(red = 2*red)

D1 <- blocks_red2 |> 
  group_by(city) |> 
  summarize(D = calcD(red, blue), .groups="drop")
```

```{r}
#| tbl-pos: H
#| tbl-cap: "Dissimilarity index before and after doubling the red population"

left_join(D0, D1, by="city", suffix=c("_orig","_red2")) |> 
  gt() |>
  fmt_number(everything(), decimals = 3)
```


Doubling the size of the red population does not change the value of the dissimilarity index.
This confirms that the dissimilarity index satisfies the composition invariance principle.

## Does the two group information theory index fulfill the composition invariance principle?
```{r}
calcH <- function(...) { # vectors or dataframe for the group counts
  t.jm<-cbind(...)      # bind together the group counts 
  t.j<-rowSums(t.jm)    # count in each neighborhood
  T<-sum(t.jm)        # total population count
  M<-ncol(t.jm)         # count the number of groups (columns)
  p.m<-colSums(t.jm)/T # calculate the population proportions
  E<-sum(p.m*log(1/p.m,M),na.rm=TRUE) # population entropy
  p.jm<-t.jm/t.j        # neighborhood-level proportions
  E.j<-rowSums(p.jm*log(1/p.jm,M),na.rm=TRUE) # neighborhood entropies
  H<-1/(T*E)*sum(t.j*(E-E.j)) # H
  return(H)
}

H0 <- blocks |> group_by(city) |> summarize(H = calcH(red, blue))
H1 <- blocks_red2 |> group_by(city) |> summarize(H = calcH(red, blue))
```

```{r}
#| tbl-pos: H
#| tbl-cap: "Information theory index (H) before vs. after doubling the red population"

left_join(H0, H1, by="city", suffix=c("_orig","_red2")) |> 
  gt() |>
  fmt_number(everything(), decimals = 3)
```

Yes. The two group information theory index fulfills the composition invariance principle.

## Demonstrate that the exposure and isolation indices are not compositionally invariant.
```{r}
expo0 <- blocks |>
  group_by(city) |>
  summarize(
    P_RR = weighted.mean(red/(red+blue),  w = red),   # isolation of red
    P_RB = weighted.mean(blue/(red+blue), w = red),   # exposure red->blue
    .groups = "drop"
  )


blocks_red2 <- blocks |> mutate(red = 2*red)

expo1 <- blocks_red2 |>
  group_by(city) |>
  summarize(
    P_RR = weighted.mean(red/(red+blue),  w = red),
    P_RB = weighted.mean(blue/(red+blue), w = red),
    .groups = "drop"
  )

expo_diff <- left_join(expo0, expo1, by="city", suffix=c("_orig","_red2")) |>
  transmute(
    City = city,
    diff_PRR = P_RR_red2 - P_RR_orig,
    diff_PRB = P_RB_red2 - P_RB_orig
  )
```
```{r}
#| tbl-pos: H
#| tbl-cap: "Change in exposure/isolation after doubling the red population"

expo_diff |> 
  gt() |>
  fmt_number(everything(), decimals = 3)
```

The table above shows that the exposure and isolation indices are not compositionally invariant.

# Experimenting with spatial segregation indices
## What happens to the segregation index as you increase or decrease the number of nearest neighbors that are considered?
```{r}
nn_bespoke<-function(g,x,y,k=10,searchtype="standard",radius=0) {
  knn<-nn2(data=cbind(x,y),k=k,searchtype=searchtype,radius=radius)
  x<-knn$nn.idx
  lapply(seq_len(nrow(x)), function(i) x[i,]) |> 
    lapply(function(x) `[`(g,x)) |> 
    lapply(table) |> 
    bind_rows() |> 
    mutate(across(everything(),as.integer))
}

k_grid <- c(5, 20)
expo_by_k <- lapply(k_grid, function(k){
  places |>
    group_by(city) |>
    mutate(nn_bespoke(group, x, y, k = k)) |>
    group_by(city, group) |>
    summarize(P_red = mean(red/k), P_blue = mean(blue/k), .groups="drop") |>
    mutate(k = k)
}) |>
  bind_rows()

expo_by_k_A <- expo_by_k |> 
  filter(city == "A")
# expo_by_k |> arrange(k, city, group) |> arrange(k, city) |> 
#   gt() |>
#   fmt_number(everything(), decimals = 3)
```

```{r}
#| tbl-pos: H

# expo_by_k_A |> 
#   filter(group=="blue") |> 
#   arrange(k, city) |> 
#   gt() |>
#   fmt_number(everything(), decimals = 3)
```


```{r}
tab_P_RR <- expo_by_k_A %>%
  filter(group == "red") %>%
  select(city, k, P_RR = P_red) %>%
  pivot_wider(names_from = k, values_from = P_RR, names_prefix = "k=") %>%
  arrange(city)

tab_P_BR <- expo_by_k_A %>%
  filter(group == "blue") %>%
  select(city, k, P_BR = P_red) %>%
  pivot_wider(names_from = k, values_from = P_BR, names_prefix = "k=") %>%
  arrange(city)
```

```{r}
#| tbl-cap: "Red isolation (P_RR) by neighborhood size (k)"

tab_P_RR |> 
  gt() |>
  fmt_number(everything(), decimals = 3)
```

```{r}
#| tbl-cap: "Blue-to-red exposure (P_BR) by neighborhood size (k)"

tab_P_BR |> 
  gt() |>
  fmt_number(everything(), decimals = 3)
```
With the number $k$ increasing, measured segregation decreases. Larger $k$ averages over broader neighborhoods, smoothing local differences.

## Does the index work as you expect for the different city scenarios? How do the values compare for City A vs. B vs. C vs. D?
```{r}
#| tbl-cap: "Blue group neighborhood composition across cities and k"

expo_by_k |> 
  filter(group=="blue") |> 
  arrange(k, city) |> 
  gt() |>
  fmt_number(everything(), decimals = 3)

# 
# tab_P_RR <- expo_by_k %>%
#   filter(group == "red") %>%
#   select(city, k, P_RR = P_red) %>%
#   pivot_wider(names_from = k, values_from = P_RR, names_prefix = "k=") %>%
#   arrange(city)
# 
# tab_P_BR <- expo_by_k %>%
#   filter(group == "blue") %>%
#   select(city, k, P_BR = P_red) %>%
#   pivot_wider(names_from = k, values_from = P_BR, names_prefix = "k=") %>%
#   arrange(city)

# tab_P_RR |> 
#   gt() |>
#   fmt_number(everything(), decimals = 3)

# tab_P_BR |> 
#   gt() |>
#   fmt_number(everything(), decimals = 3)
```

Yes. Across all neighborhood sizes, segregation is highest in City A, followed by B, C, and D.
This ordering matches the increasing level of noise introduced in each scenario.

## How do the values of the spatial dissimilarity index vary across the different cities as you increase or decrease the number of nearest neighbors?

```{r}
calcD<-function(g,...) {
  p.j<-cbind(...)
  p.j<-p.j/rowSums(p.j)
  p<-table(g)
  T<-sum(p)
  p<-p/T
  nms<-names(p)
  result<-1/(2*T)*sum(abs(p.j[,nms[1]]/p[1] - p.j[,nms[2]]/p[2]))
  return(result)
}

spatial_dissimilarity <- function(city_name = "A", k_num = 10) {
  city<-filter(places,city==city_name)
  knn_nbhd<-nn2(data=city[,c("x","y")],k=k_num)

  x<-knn_nbhd$nn.idx
  x<-lapply(seq_len(nrow(x)), function(i) x[i,])

  g<-pull(city,group)
  x<-lapply(x,function(x) `[`(g,x)) 
  bespoke<-lapply(x,table) |> 
      bind_rows() |> 
     mutate(across(everything(),as.integer)) 
  bespoke_city<-bind_cols(city,bespoke)
  bespoke_city |> summarize(D=calcD(group,red,blue))
}
```

```{r}
cities <- c("A","B","C","D")
ks <- c(5, 10, 20)
grid <- tidyr::expand_grid(city = cities, k = ks)

sd_all <- grid |>
  rowwise() |>
  mutate(D = spatial_dissimilarity(city, k)$D) |>
  ungroup()
```

```{r}
#| fig-cap: "Spatial dissimilarity of Each City"

ggplot(sd_all, aes(x = k, y = D, group = city)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(trans = "log10") +
  labs(x = "k (nearest neighbors)", y = "Spatial dissimilarity (D)")
```

For all cities, spatial dissimilarity declines as k increases.
Cities with stronger underlying spatial structure (A and B) maintain higher dissimilarity at all scales.
Cities with more randomness (C and D) show lower values and converge more quickly as neighborhoods grow.
