---
title: "Inequality and Segregation Lab 3"
author: "Yukun Jiao"
date: "2025-12-01"
toc: false
format:
  html: 
    embed-resources: true
knitr: 
  opts_chunk:
    fig.align: center
editor: 
  markdown: 
    wrap: 72
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(ggforce)
library(cowplot)
library(RANN)
library(sf)
library(gt)
library(knitr)
library(kableExtra)

make_housing<-function(dim=100,n=10000,foci=c(0.2,0.8),noise=1,pow=2) {
  xy <-bind_rows(
    expand_grid( x=seq(0,dim)/dim, y=seq(0.5,dim-0.5)/dim ) ,
    expand_grid( x=seq(0.5,dim-0.5)/dim , y=seq(0,dim)/dim) )
  if(n>=nrow(xy)) {
    return(xy)
  }
  else {
    foci<-as_tibble(expand.grid(x_f=foci,y_f=foci))
    places<-expand_grid(xy,foci) |> 
      mutate(d=sqrt((x-x_f)^2+(y-y_f)^2)) |> 
      group_by(x,y) |> 
      summarize(u=sum(1/(d+0.001)^pow)) |> 
      ungroup() |> 
      mutate(u=u/sd(u)) |> 
      mutate(v = u + rnorm(n(),0,sd=noise)) |> 
      filter(v>=sort(v,decreasing=T)[n]) |> 
      select(x,y)
    return(places)
  }
}
```


# Modifying the Segregation Scenarios

## reducing to a single focus.
```{r}
#| warning: false

foci<-as_tibble(expand.grid(x_f=c(0.5),y_f=c(0.5)))


places<-make_housing(pow=1,noise=0.5,foci=c(0.5),n=10000)
# ggplot() +
#   geom_point(data=places,aes(x=x,y=y),size=0.1) +
#   geom_point(data=foci,aes(x=x_f,y=y_f),color="green3",size=5,shape=4,stroke=3) +
#   theme_void() + coord_fixed()

places<-places |> 
  expand_grid(foci) |> 
  mutate(d=sqrt((x-x_f)^2+(y-y_f)^2)) |> 
  mutate(u=1/d) |> 
  group_by(x,y) |> 
  summarize(u=sum(u)) |> 
  ungroup() |> # do this so calculations aren't performed within combos of x & y
  mutate(C=u+rnorm(n=length(u),mean=0,sd=sd(u))) |> 
  mutate(C=factor(ifelse(C>median(C),"red","blue"))) |> 
  ungroup() |> 
  mutate(A=u+rnorm(n=length(u),mean=0,sd=0*sd(u))) |>  # no noise!
  mutate(A=factor(ifelse(A>median(A),"red","blue"))) |> 
  mutate(B=u+rnorm(n=length(u),mean=0,sd=0.2*sd(u))) |> # a bit of noise
  mutate(B=factor(ifelse(B>median(B),"red","blue"))) |> 
  mutate(D=u+rnorm(n=length(u),mean=0,sd=3*sd(u))) |> # a lot of noise
  mutate(D=factor(ifelse(D>median(D),"red","blue")))

# ggplot(places,aes(x=x,y=y,color=C)) +
#   geom_point() + 
#   scale_color_identity() +
#   theme_void() + 
#   coord_fixed() + 
#   theme(legend.position = "none")

places <- places |> pivot_longer(cols=c("A","B","C","D"),
               names_to="city",
               values_to="group")

ggplot(data=slice(places,sample(1:nrow(places))), # random plot order
       aes(x=x,y=y,color=group)) + 
  geom_point(size=0.1) + 
  scale_color_identity() +
  coord_fixed() +
  theme_void() + 
  theme(legend.position = "none", aspect.ratio = 1) + 
  facet_grid(cols=vars(city)) -> gg_points

# gg_points

grid_dim<-10
blocks<-places |> 
  mutate(x=ceiling(x*grid_dim)/grid_dim,y=ceiling(y*grid_dim)/grid_dim) |> 
  group_by(city,x,y) |> 
  summarize(red=sum(ifelse(group=="red",1,0)),
            blue=sum(ifelse(group=="blue",1,0))) |> 
  mutate(`% red`=100*red/(red+blue))

ggplot(blocks,aes(x=x,y=y,fill=`% red`)) + 
  geom_raster() + 
  facet_grid(cols=vars(city)) + 
  coord_fixed() +
  scale_fill_gradient2(midpoint=50,high="red",mid="white",low="blue") + 
  theme_void() ->
  gg_blocks

plot_grid(gg_points,gg_blocks,nrow=2,align="hv",axis="r")
```

## Increasing the number of foci.
```{r}
#| warning: false

foci<-as_tibble(expand.grid(x_f=c(0.1, 0.5, 0.75),y_f=c(0.1, 0.5, 0.75)))

places<-make_housing(pow=1,noise=0.5,foci=c(0.5),n=10000)
# ggplot() +
#   geom_point(data=places,aes(x=x,y=y),size=0.1) +
#   geom_point(data=foci,aes(x=x_f,y=y_f),color="green3",size=5,shape=4,stroke=3) +
#   theme_void() + coord_fixed()

places<-places |> 
  expand_grid(foci) |> 
  mutate(d=sqrt((x-x_f)^2+(y-y_f)^2)) |> 
  mutate(u=1/d) |> 
  group_by(x,y) |> 
  summarize(u=sum(u)) |> 
  ungroup() |> # do this so calculations aren't performed within combos of x & y
  mutate(C=u+rnorm(n=length(u),mean=0,sd=sd(u))) |> 
  mutate(C=factor(ifelse(C>median(C),"red","blue"))) |> 
  ungroup() |> 
  mutate(A=u+rnorm(n=length(u),mean=0,sd=0*sd(u))) |>  # no noise!
  mutate(A=factor(ifelse(A>median(A),"red","blue"))) |> 
  mutate(B=u+rnorm(n=length(u),mean=0,sd=0.2*sd(u))) |> # a bit of noise
  mutate(B=factor(ifelse(B>median(B),"red","blue"))) |> 
  mutate(D=u+rnorm(n=length(u),mean=0,sd=3*sd(u))) |> # a lot of noise
  mutate(D=factor(ifelse(D>median(D),"red","blue")))


# ggplot(places,aes(x=x,y=y,color=C)) +
#   geom_point() + 
#   scale_color_identity() +
#   theme_void() + 
#   coord_fixed() + 
#   theme(legend.position = "none")

places <- places |> pivot_longer(cols=c("A","B","C","D"),
               names_to="city",
               values_to="group")

ggplot(data=slice(places,sample(1:nrow(places))), # random plot order
       aes(x=x,y=y,color=group)) + 
  geom_point(size=0.1) + 
  scale_color_identity() +
  coord_fixed() +
  theme_void() + 
  theme(legend.position = "none", aspect.ratio = 1) + 
  facet_grid(cols=vars(city)) -> gg_points

# gg_points


grid_dim<-10
blocks<-places |> 
  mutate(x=ceiling(x*grid_dim)/grid_dim,y=ceiling(y*grid_dim)/grid_dim) |> 
  group_by(city,x,y) |> 
  summarize(red=sum(ifelse(group=="red",1,0)),
            blue=sum(ifelse(group=="blue",1,0))) |> 
  mutate(`% red`=100*red/(red+blue))

ggplot(blocks,aes(x=x,y=y,fill=`% red`)) + 
  geom_raster() + 
  facet_grid(cols=vars(city)) + 
  coord_fixed() +
  scale_fill_gradient2(midpoint=50,high="red",mid="white",low="blue") + 
  theme_void() ->
  gg_blocks

plot_grid(gg_points,gg_blocks,nrow=2,align="hv",axis="r")
```
## Changing the spatial distribution of the foci.
```{r}
#| warning: false

foci<-as_tibble(expand.grid(x_f=c(0.1),y_f=c(0.1)))

places<-make_housing(pow=1,noise=0.5,foci=c(0.5),n=10000)
ggplot() +
  geom_point(data=places,aes(x=x,y=y),size=0.1) +
  geom_point(data=foci,aes(x=x_f,y=y_f),color="green3",size=5,shape=4,stroke=3) +
  theme_void() + coord_fixed()

places<-places |> 
  expand_grid(foci) |> 
  mutate(d=sqrt((x-x_f)^2+(y-y_f)^2)) |> 
  mutate(u=1/d) |> 
  group_by(x,y) |> 
  summarize(u=sum(u)) |> 
  ungroup() |> # do this so calculations aren't performed within combos of x & y
  mutate(C=u+rnorm(n=length(u),mean=0,sd=sd(u))) |> 
  mutate(C=factor(ifelse(C>median(C),"red","blue"))) |> 
  ungroup() |> 
  mutate(A=u+rnorm(n=length(u),mean=0,sd=0*sd(u))) |>  # no noise!
  mutate(A=factor(ifelse(A>median(A),"red","blue"))) |> 
  mutate(B=u+rnorm(n=length(u),mean=0,sd=0.2*sd(u))) |> # a bit of noise
  mutate(B=factor(ifelse(B>median(B),"red","blue"))) |> 
  mutate(D=u+rnorm(n=length(u),mean=0,sd=3*sd(u))) |> # a lot of noise
  mutate(D=factor(ifelse(D>median(D),"red","blue")))


# ggplot(places,aes(x=x,y=y,color=C)) +
#   geom_point() + 
#   scale_color_identity() +
#   theme_void() + 
#   coord_fixed() + 
#   theme(legend.position = "none")

places <- places |> pivot_longer(cols=c("A","B","C","D"),
               names_to="city",
               values_to="group")

ggplot(data=slice(places,sample(1:nrow(places))), # random plot order
       aes(x=x,y=y,color=group)) + 
  geom_point(size=0.1) + 
  scale_color_identity() +
  coord_fixed() +
  theme_void() + 
  theme(legend.position = "none", aspect.ratio = 1) + 
  facet_grid(cols=vars(city)) -> gg_points
gg_points
```


<!-- -   Switching to a linear assignment function. -->
<!-- -   Applying a logarithmic or some other non-linear assignment function. -->
<!-- -   Increasing the share of red dots. -->
<!-- -   Decreasing the share of red dots. -->
<!-- -   CHALLENGE: Making the *locations* (*not* a particular group) denser around the foci. -->




# Evenness and exposure measures of segregation

## Confirm that the index of dissimilarity obeys composition invariance principle by doubling the size of the "red" population in each neighborhood and re-calculating the indices.
```{r}
calcD <- function(x,y) {
  result<-0.5*sum(abs(x/sum(x)-y/sum(y)))
  return(result)
}

grid_dim<-10
blocks<-places |> 
  mutate(x=ceiling(x*grid_dim)/grid_dim,y=ceiling(y*grid_dim)/grid_dim) |> 
  group_by(city,x,y) |> 
  summarize(red=sum(ifelse(group=="red",1,0)),
            blue=sum(ifelse(group=="blue",1,0))) |> 
  mutate(`% red`=100*red/(red+blue))


D0 <- blocks |> 
  group_by(city) |> 
  summarize(D = calcD(red, blue), .groups="drop")

blocks_red2 <- blocks |> mutate(red = 2*red)

D1 <- blocks_red2 |> 
  group_by(city) |> 
  summarize(D = calcD(red, blue), .groups="drop")
```

```{r}
#| tbl-pos: H

left_join(D0, D1, by="city", suffix=c("_orig","_red2")) |> 
  gt() |>
  fmt_number(everything(), decimals = 3)
```



## Does the two group information theory index fulfill the composition invariance principle?
```{r}
calcH <- function(...) { # vectors or dataframe for the group counts
  t.jm<-cbind(...)      # bind together the group counts 
  t.j<-rowSums(t.jm)    # count in each neighborhood
  T<-sum(t.jm)        # total population count
  M<-ncol(t.jm)         # count the number of groups (columns)
  p.m<-colSums(t.jm)/T # calculate the population proportions
  E<-sum(p.m*log(1/p.m,M),na.rm=TRUE) # population entropy
  p.jm<-t.jm/t.j        # neighborhood-level proportions
  E.j<-rowSums(p.jm*log(1/p.jm,M),na.rm=TRUE) # neighborhood entropies
  H<-1/(T*E)*sum(t.j*(E-E.j)) # H
  return(H)
}

H0 <- blocks |> group_by(city) |> summarize(H = calcH(red, blue))
H1 <- blocks_red2 |> group_by(city) |> summarize(H = calcH(red, blue))

left_join(H0, H1, by="city", suffix=c("_orig","_red2")) |> 
  gt() |>
  fmt_number(everything(), decimals = 3)
```

```{r}
#| tbl-pos: H

left_join(H0, H1, by="city", suffix=c("_orig","_red2")) |> 
  gt() |>
  fmt_number(everything(), decimals = 3)
```


## Demonstrate that the exposure and isolation indices are not compositionally invariant.
```{r}
expo0 <- blocks |>
  group_by(city) |>
  summarize(
    P_RR = weighted.mean(red/(red+blue),  w = red),   # isolation of red
    P_RB = weighted.mean(blue/(red+blue), w = red),   # exposure red->blue
    .groups = "drop"
  )


blocks_red2 <- blocks |> mutate(red = 2*red)

expo1 <- blocks_red2 |>
  group_by(city) |>
  summarize(
    P_RR = weighted.mean(red/(red+blue),  w = red),
    P_RB = weighted.mean(blue/(red+blue), w = red),
    .groups = "drop"
  )

expo_diff <- left_join(expo0, expo1, by="city", suffix=c("_orig","_red2")) |>
  transmute(
    City = city,
    diff_PRR = P_RR_red2 - P_RR_orig,
    diff_PRB = P_RB_red2 - P_RB_orig
  )
```
```{r}
#| tbl-pos: H

expo_diff |> 
  gt() |>
  fmt_number(everything(), decimals = 3)
```


<!-- -   What happens to your measures of evenness as you divide up the city into more or fewer blocks? -->
<!-- -   How low do the segregation indices go when agents are assigned totally at random (i.e., no relationship between distance to foci and placement of groups, or lots of random noise). -->


# Experimenting with spatial segregation indices
## What happens to the segregation index as you increase or decrease the number of nearest neighbors that are considered?
```{r}
nn_bespoke<-function(g,x,y,k=10,searchtype="standard",radius=0) {
  knn<-nn2(data=cbind(x,y),k=k,searchtype=searchtype,radius=radius)
  x<-knn$nn.idx
  lapply(seq_len(nrow(x)), function(i) x[i,]) |> 
    lapply(function(x) `[`(g,x)) |> 
    lapply(table) |> 
    bind_rows() |> 
    mutate(across(everything(),as.integer))
}

k_grid <- c(5, 20)
expo_by_k <- lapply(k_grid, function(k){
  places |>
    group_by(city) |>
    mutate(nn_bespoke(group, x, y, k = k)) |>
    group_by(city, group) |>
    summarize(P_red = mean(red/k), P_blue = mean(blue/k), .groups="drop") |>
    mutate(k = k)
}) |>
  bind_rows()

# expo_by_k |> arrange(k, city, group) |> arrange(k, city) |> 
#   gt() |>
#   fmt_number(everything(), decimals = 3)


```
```{r}
#| tbl-pos: H

expo_by_k |> filter(group=="blue") |> arrange(k, city) |> 
  gt() |>
  fmt_number(everything(), decimals = 3)
```


```{r}
tab_P_RR <- expo_by_k %>%
  filter(group == "red") %>%
  select(city, k, P_RR = P_red) %>%
  pivot_wider(names_from = k, values_from = P_RR, names_prefix = "k=") %>%
  arrange(city)

tab_P_BR <- expo_by_k %>%
  filter(group == "blue") %>%
  select(city, k, P_BR = P_red) %>%
  pivot_wider(names_from = k, values_from = P_BR, names_prefix = "k=") %>%
  arrange(city)
```

```{r}
#| tbl-cap: "Red isolation (P_RR) by neighborhood size (k)"

tab_P_RR |> 
  gt() |>
  fmt_number(everything(), decimals = 3)
```

```{r}
#| tbl-cap: "Blue-to-red exposure (P_BR) by neighborhood size (k)"

tab_P_BR |> 
  gt() |>
  fmt_number(everything(), decimals = 3)
```

